% =============================================================================
% The Mathematics of Backpropagation
%
% Companion notes to llm.mojo: one section per op in the GPT-2 forward pass,
% in llm.c order. Each section states the forward computation (the anchor),
% then leaves a Derivation block to work the backward pass by hand, and a
% Result block to record the final gradient(s) once derived.
%
% Build:  make docs                      (from the repo root)
%         latexmk -pdf backprop.tex      (from docs/, equivalent)
% =============================================================================
\documentclass[letterpaper, 10 pt, conference]{ieeeconf}

\let\proof\relax
\let\endproof\relax

\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage[colorlinks=true, linkcolor=blue!50!black]{hyperref}

\IEEEoverridecommandlockouts
\overrideIEEEmargins

% -----------------------------------------------------------------------------
% Macros
% -----------------------------------------------------------------------------
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}

% Partial and total derivatives: \pdv{f}{x}, \dv{f}{x}
\newcommand{\pdv}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\dv}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}

% Upstream gradient of the loss w.r.t. a tensor: \dL{x} renders dL/dx.
% This is the quantity every backward kernel receives or produces.
\newcommand{\dL}[1]{\pdv{L}{#1}}

% One-hot indicator and Kronecker delta
\newcommand{\ind}[1]{\mathbb{1}\!\left[#1\right]}
\newcommand{\kron}[2]{\delta_{#1 #2}}

% softmax / layernorm / gelu as upright operators
\DeclareMathOperator{\softmax}{softmax}
\DeclareMathOperator{\layernorm}{layernorm}
\DeclareMathOperator{\gelu}{GELU}
\DeclareMathOperator{\erf}{erf}

% Workspace marker for unfinished derivations
\newcommand{\TODO}[1]{\textcolor{red!70!black}{\textbf{TODO:} #1}}

% Derivation / Result blocks
\theoremstyle{definition}
\newtheorem*{derivation}{Derivation}
\newtheorem*{result}{Result}

\title{\LARGE \bf
The Mathematics of Backpropagation:\\
Companion Derivations for \texttt{llm.mojo}
}
\author{John W. Owen III}

% =============================================================================
\begin{document}
\maketitle

\begin{abstract}
  This document presents working derivations of the backward pass for every
  operation in the GPT-2 forward pass, as implemented in the \texttt{llm.mojo}
  kernels. The project is directly inspired by Andrej Karpathy's \texttt{llm.c}~\cite{llmc}.
  We order the sections by backward-derivation pedagogy, starting from the scalar loss
  and working toward the inputs, rather than by the forward execution order of \texttt{llm.c}.
  This structure ensures that we derive each gradient only after deriving the gradient on
  which it depends. Each section anchors on the forward computation and provides a structured
  workspace for the derivation. The companion source code is available at \url{https://github.com/ulmentflam/llm.mojo}.
\end{abstract}

\tableofcontents

% =============================================================================
\section{Notation and Conventions}\label{sec:notation}

Tensor shapes in these derivations follow the parameters of the implementation, which Table~\ref{tab:notation} summarizes. We define the index conventions as $b$ for the batch, $t$ for sequence positions, and $i, j, k$ for channels or vocabulary entries. A row of activations at a single sequence position is denoted by the vector $x_{bt} \in \R^{C}$.

\begin{table}[t]
  \centering
  \caption{Symbols and their corresponding kernel parameters.}
  \label{tab:notation}
  \begin{tabular}{@{}lll@{}}
    \toprule
    Symbol & Code                         & Meaning                       \\
    \midrule
    $B$    & \texttt{batch\_size}         & batch size                    \\
    $T$    & \texttt{seq\_len}            & sequence length               \\
    $C$    & \texttt{channels}            & model width (embed dim)       \\
    $V$    & \texttt{vocab\_size}         & real vocabulary size          \\
    $V_p$  & \texttt{vocab\_size\_padded} & padded vocab (stride)         \\
    $H$    & \texttt{num\_heads}          & attention heads, head $h=C/H$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Why index form?}
Matrix calculus obscures backpropagation under transposes and Kronecker products. This document writes every relation in \emph{index form} (element-by-element), reducing tensor operations to scalar calculus. Treating entries as independent scalar variables allows standard derivative rules. Matching indices then reassembles scalar equations back into matrix form.

\paragraph{Linear algebra in index form.}
A matrix entry is named (row, column): $A_{ij}$ is row $i$, column $j$. Three definitions cover every matrix manipulation in this document. The transpose swaps index roles:
\begin{equation}
  (A^{\top})_{ij} \;=\; A_{ji}.
  \label{eq:transpose}
\end{equation}
The matrix product contracts the inner indices:
\begin{equation}
  (AB)_{ik} \;=\; \sum_j A_{ij}\,B_{jk}.
  \label{eq:matprod}
\end{equation}
In code, this summation represents a loop over the shared dimension $j$:
\begin{verbatim}
val = 0.0
for j in range(width):
    val += A[i, j] * B[j, k]
\end{verbatim}
Reversing \eqref{eq:matprod} identifies matrix products. Summing over a shared index yields a product when the index is the left factor's column and the right factor's row. Elsewhere, a transpose \eqref{eq:transpose} restores this alignment. This generates every transpose in the backward pass.
Adding a vector to a matrix broadcasts the vector over rows:
\begin{equation}
  (M + b)_{ni} \;=\; M_{ni} + b_i.
  \label{eq:broadcast}
\end{equation}

\paragraph{Target and probe indices.}
Differentiating a tensor expression requires distinct names for the output (\emph{target}) indices and the differentiation (\emph{probe}) indices. Perturbing $X_{mk}$ to find the change in $Y_{ni}$ requires independent letters ($m, k \neq n, i$). Reusing indices would restrict the perturbation to matching coordinate pairs. Independent probe indices evaluate the derivative of any output coordinate with respect to any input coordinate.

\paragraph{Indexed variables under differentiation.}
Entries of distinct tensors are unrelated; their cross-derivatives vanish. The derivative of a tensor entry with respect to another is one for identical coordinates and zero otherwise. The Kronecker delta records this relation:
\begin{equation}
  \kron{i}{j} \;=\;
  \begin{cases}
    1 & i = j,    \\
    0 & i \neq j.
  \end{cases}
  \label{eq:kron}
\end{equation}
For a vector, one index pair determines equality; for a matrix, both row and column must match. A product of deltas encodes this conjunction:
\begin{equation}
  \pdv{x_j}{x_i} \;=\; \kron{j}{i},
  \qquad
  \pdv{W_{ij}}{W_{kl}} \;=\; \kron{i}{k}\,\kron{j}{l}.
  \label{eq:entry-dv}
\end{equation}
Two concrete matrix instances:
\begin{equation}
  \pdv{W_{12}}{W_{12}} = \kron{1}{1}\,\kron{2}{2} = 1 \cdot 1 = 1,
  \quad
  \pdv{W_{12}}{W_{21}} = \kron{1}{2}\,\kron{2}{1} = 0 \cdot 0 = 0.
  \label{eq:entry-dv-example}
\end{equation}
Deltas exit derivations by collapsing sums. For fixed $l$, the Kronecker delta sifts the summation:
\begin{align}
  \sum_{j} \kron{j}{l}\,A_j \;=\; A_l.
  \label{eq:sift}
\end{align}
This sifting acts as a conditional check inside a loop:
\begin{verbatim}
sum_val = 0.0
for j in range(N):
    if j == l:
        sum_val += 1.0 * A[j]
    else:
        sum_val += 0.0 * A[j]
# sum_val is now exactly A[l]
\end{verbatim}
Unsummed deltas remain in the final expression, enforcing index equality.

\paragraph{The backward convention.}
Each operation receives the upstream gradient $\dL{y}$ of the scalar loss $L$ and computes the downstream gradient $\dL{x}$ (or parameter gradient $\dL{w}$) via the chain rule:
\begin{equation}
  \dL{x_i} \;=\; \sum_j \dL{y_j}\,\pdv{y_j}{x_i}.
  \label{eq:chain}
\end{equation}
This sum accumulates gradients from all output coordinates $y_j$ that depend on $x_i$. Shared inputs or parameters (such as tied weights or broadcast biases) accumulate gradients across all active positions via in-place addition ($\mathrel{+}=$).

\paragraph{Layout and transposes.}
The \emph{numerator layout} convention matches the shape of the gradient $\dL{X}$ to the source tensor $X$. Index-level rearrangements then generate transposes automatically, avoiding memorized matrix identities.

\paragraph{Reading order.}
While the forward pass runs encoder $\to$ layernorm $\to$ attention $\to$ matmul $\to$ GELU $\to$ softmax $\to$ cross-entropy, these derivations follow the backward flow. Starting at the loss (\S\ref{sec:crossentropy}) and ending at the embeddings (\S\ref{sec:encoder}), each section consumes the upstream gradient from the previous section. \S\ref{sec:full} reassembles the complete backward pass.


% =============================================================================
\section{Scalar Example: Chain Rule on a Tiny Network}\label{sec:example}

We introduce a two-step composition to fix conventions before the tensor operations:
$z = wx + b$, $a = \sigma(z)$, and $L = \tfrac{1}{2}(a - y)^2$.

\begin{derivation}
  We propagate the gradient backward through the loss, then the sigmoid:
  \begin{align}
    \dL{a}     & = a - y                                                              \\
    \dL{z}     & = \dL{a} \cdot \pdv{a}{z}                                            \\
    \pdv{a}{z} & = \sigma(z) \cdot (1-\sigma(z))
    \intertext{Because $a = \sigma(z)$,}
    \dL{z}     & = (a - y) \cdot \pdv{a}{z} = (a - y) \cdot a \cdot (1-a).
    \intertext{We compute the gradient for the weight:}
    \dL{w}     & = \dL{a} \cdot \pdv{a}{z} \cdot \pdv{z}{w} = \dL{z} \cdot \pdv{z}{w} \\
    \pdv{z}{w} & = x                                                                  \\
    \dL{w}     & = \dL{z} \cdot x = (a - y) \cdot a \cdot (1-a) \cdot x.
    \intertext{We compute the gradient for the bias:}
    \dL{b}     & = \dL{a} \cdot \pdv{a}{z} \cdot \pdv{z}{b} = \dL{z} \cdot \pdv{z}{b} \\
    \pdv{z}{b} & = 1                                                                  \\
    \dL{b}     & = \dL{z} = (a - y) \cdot a \cdot (1-a).
    \intertext{We compute the gradient for the input:}
    \dL{x}     & = \dL{a} \cdot \pdv{a}{z} \cdot \pdv{z}{x} = \dL{z} \cdot \pdv{z}{x} \\
    \pdv{z}{x} & = w                                                                  \\
    \dL{x}     & = \dL{z} \cdot w = (a - y) \cdot a \cdot (1-a) \cdot w.
  \end{align}
\end{derivation}

\begin{result}
  The gradient propagates in a single chain back through the network, fanning out at the affine inputs:
  \begin{equation}
    \dL{a} \;\to\; \dL{z} \;\to\; \Bigl\{\dL{w},\; \dL{b},\; \dL{x}\Bigr\}
  \end{equation}
  where we compute and reuse $\dL{z} = (a - y) \cdot a \cdot (1-a)$:
  $\dL{w} = \dL{z} \cdot x$, \quad $\dL{b} = \dL{z}$, \quad
  $\dL{x} = \dL{z} \cdot w$.
\end{result}

% =============================================================================
\section{Cross-Entropy Loss}\label{sec:crossentropy}

We define the standard cross-entropy between a target distribution $y_{bt} \in \R^{V}$ and predicted probabilities $p_{bt} \in \R^{V_p}$ as:
\begin{equation}
  L_{bt} \;=\; -\sum_{i<V} y_{bt,i} \log p_{bt,i}.
  \label{eq:ce-std}
\end{equation}
For language modeling, the target distribution is one-hot. Writing $\ind{P}$ for the indicator of a condition $P$:
\begin{equation}
  \ind{P} \;=\;
  \begin{cases}
    1 & P \text{ is true},  \\
    0 & P \text{ is false},
  \end{cases}
  \label{eq:indicator}
\end{equation}
(where the Kronecker delta \eqref{eq:kron} is this indicator specialized to index equality, $\kron{i}{j} = \ind{i = j}$), the target with token $y_{bt} \in \{0,\dots,V-1\}$ is $y_{bt,i} = \ind{i = y_{bt}}$, placing all probability mass on the target token and zero elsewhere. Every term of \eqref{eq:ce-std} except one vanishes, collapsing the loss to the form that \texttt{crossentropy\_ohe\_fwd} computes:
\begin{equation}
  L_{bt} \;=\; -\log p_{bt,\,y_{bt}}.
  \label{eq:ce-fwd}
\end{equation}

\subsection{Backward w.r.t.\ the probabilities}

\begin{derivation}
  We assign index $i$ to the probed coordinate of $p_{bt}$ and index $j$ to the summed coordinate. Because rows are independent, the coordinates of $p_{bt}$ within a row act as independent variables obeying \eqref{eq:entry-dv}. We expand the standard form \eqref{eq:ce-std} term by term:
  \begin{multline}
    L_{bt} = -\bigl(
    y_{bt,0}\log p_{bt,0}
    + y_{bt,1}\log p_{bt,1} \\
    + y_{bt,2}\log p_{bt,2}
    + \cdots
    + y_{bt,V-1}\log p_{bt,V-1}
    \bigr).
    \label{eq:ce-std-expanded}
  \end{multline}
  When we differentiate with respect to one coordinate $p_{bt,i}$, every term of the expanded sum except the $i$-th term is constant, so only its derivative survives:
  \begin{align}
    \dL{p_{bt,i}}
     & = -\pdv{}{p_{bt,i}}\bigl(
    y_{bt,0}\log p_{bt,0} + \cdots
    + y_{bt,i}\log p_{bt,i} \notag                   \\
     & \qquad\qquad + \cdots
    + y_{bt,V-1}\log p_{bt,V-1}
    \bigr) \notag                                    \\
     & = -\,y_{bt,i}\,\pdv{}{p_{bt,i}}\log p_{bt,i}.
    \label{eq:ce-surviving}
  \end{align}
  The surviving factor is the derivative of the logarithm:
  \begin{equation}
    \dv{}{x}\log x \;=\; \frac{1}{x}
    \qquad\Longrightarrow\qquad
    \pdv{}{p_{bt,i}}\log p_{bt,i} \;=\; \frac{1}{p_{bt,i}}.
    \label{eq:dlog}
  \end{equation}
  In delta form, we apply \eqref{eq:entry-dv} followed by the sifting property \eqref{eq:sift}: each term of the sum differentiates to $(y_{bt,j}/p_{bt,j})\,\kron{j}{i}$, which collapses the sum to the single term where $j = i$:
  \begin{equation}
    \dL{p_{bt,i}}
    \;=\;
    \sum_{j<V} -\frac{y_{bt,j}}{p_{bt,j}}\,\kron{j}{i}
    \;=\;
    -\frac{y_{bt,i}}{p_{bt,i}},
    \label{eq:ce-delta-form}
  \end{equation}
  yielding the standard-sum gradient, which remains valid for \emph{any} target distribution $y_{bt}$:
  \begin{equation}
    \dL{p_{bt,i}}
    \;=\;
    -\,\pdv{}{p_{bt,i}} \sum_{j<V} y_{bt,j} \log p_{bt,j}
    \;=\;
    -\frac{y_{bt,i}}{p_{bt,i}}.
    \label{eq:ce-std-bwd}
  \end{equation}
  When we substitute the one-hot target $y_{bt,i} = \ind{i = y_{bt}}$, the loss \eqref{eq:ce-fwd} depends only on the target coordinate $p_{bt,\,y_{bt}}$:
  \begin{equation}
    \dL{p_{bt,i}} =
    \begin{cases}
      -\dfrac{1}{p_{bt,\,y_{bt}}} & i = y_{bt},       \\
      0                           & \text{otherwise}.
    \end{cases}
  \end{equation}
\end{derivation}

\begin{result}
  \begin{equation}
    \dL{p_{bt,i}} \;=\; -\frac{\ind{i = y_{bt}}}{p_{bt,i}}.
    \label{eq:ce-bwd}
  \end{equation}
\end{result}

% =============================================================================
\section{Softmax}\label{sec:softmax}

We express the forward pass from logits $\ell_{bt} \in \R^{V_p}$ with real entries $i < V$ as:
\begin{equation}
  p_{bt,i} \;=\; \frac{e^{\ell_{bt,i} - m_{bt}}}{\sum_{j<V} e^{\ell_{bt,j} - m_{bt}}},
  \qquad m_{bt} = \max_{j<V} \ell_{bt,j}.
  \label{eq:softmax-fwd}
\end{equation}

\subsection{The Jacobian}

We drop the row indices $bt$ throughout this section because each row represents an independent softmax, making the Jacobian per-row.

\begin{derivation}
  We seek the Jacobian entry $\pdv{p_i}{\ell_j}$, which describes how one probability changes with respect to one logit. We assign index $i$ to the probed probability and index $j$ to the probed logit. The logits within a row are independent variables obeying \eqref{eq:entry-dv}, $\pdv{\ell_a}{\ell_j} = \kron{a}{j}$. The loss enters only in the subsequent sections, where we chain an upstream gradient through these entries. Before differentiating, we note that the max-subtraction in \eqref{eq:softmax-fwd} cancels because for any constant $c$:
  \begin{equation}
    \frac{e^{\ell_i - c}}{\sum_{j<V} e^{\ell_j - c}}
    \;=\;
    \frac{e^{-c}\,e^{\ell_i}}{e^{-c}\sum_{j<V} e^{\ell_j}}
    \;=\;
    \frac{e^{\ell_i}}{\sum_{j<V} e^{\ell_j}},
    \label{eq:softmax-shift}
  \end{equation}
  allowing us to differentiate the unshifted form. We expand the denominator sum in the same format:
  \begin{equation}
    p_i = \frac{e^{\ell_i}}{S},
    \qquad
    S = e^{\ell_0} + e^{\ell_1} + e^{\ell_2} + \cdots + e^{\ell_{V-1}}.
    \label{eq:softmax-S}
  \end{equation}
  We compute each entry using the quotient rule for a fraction $u/v$ with scalar argument $x$:
  \begin{equation}
    \pdv{}{x}\,\frac{u}{v}
    \;=\;
    \frac{\pdv{u}{x}\,v \;-\; u\,\pdv{v}{x}}{v^{2}}.
    \label{eq:quotient}
  \end{equation}
  We calculate the two-by-two block of entries for $p_1, p_2$ and $\ell_1, \ell_2$ explicitly before generalizing.

  \paragraph{Entry $\pdv{p_1}{\ell_1}$}
  In this case, $u = e^{\ell_1}$ and $v = S$, and both variables vary with $\ell_1$:
  \begin{align}
    \pdv{u}{\ell_1} & = \pdv{}{\ell_1}\,e^{\ell_1} = e^{\ell_1},
    \label{eq:q11-u}                                                \\
    \pdv{v}{\ell_1} & = \pdv{}{\ell_1}\bigl(e^{\ell_0} + e^{\ell_1}
    + e^{\ell_2} + \cdots + e^{\ell_{V-1}}\bigr) = e^{\ell_1},
    \label{eq:q11-v}
  \end{align}
  where every term of the expanded sum except $e^{\ell_1}$ is constant. Substituting \eqref{eq:q11-u} and \eqref{eq:q11-v} into the quotient rule \eqref{eq:quotient} yields:
  \begin{align}
    \pdv{p_1}{\ell_1}
     & = \frac{e^{\ell_1}\bigl(e^{\ell_0} + e^{\ell_1} + \cdots
      + e^{\ell_{V-1}}\bigr)
      - e^{\ell_1}\,e^{\ell_1}}
    {\bigl(e^{\ell_0} + e^{\ell_1} + \cdots + e^{\ell_{V-1}}\bigr)^{2}}
    \notag                                                      \\
     & = \frac{e^{\ell_1}\,S - e^{\ell_1}\,e^{\ell_1}}{S^{2}}
    = \frac{e^{\ell_1}}{S}
    - \left(\frac{e^{\ell_1}}{S}\right)^{\!2} \notag            \\
     & = p_1 - p_1^{2} \;=\; p_1\,(1 - p_1).
    \label{eq:q11}
  \end{align}

  \paragraph{Entry $\pdv{p_1}{\ell_2}$}
  Here, the numerator does not change: $u = e^{\ell_1}$ is constant with respect to $\ell_2$, which gives $\pdv{u}{\ell_2} = 0$. Again, only one term of the expanded denominator sum survives:
  \begin{equation}
    \pdv{v}{\ell_2} = \pdv{}{\ell_2}\bigl(e^{\ell_0} + e^{\ell_1}
    + e^{\ell_2} + \cdots + e^{\ell_{V-1}}\bigr) = e^{\ell_2}.
    \label{eq:q12-v}
  \end{equation}
  Substituting these values into \eqref{eq:quotient} yields:
  \begin{align}
    \pdv{p_1}{\ell_2}
     & = \frac{0 \cdot \bigl(e^{\ell_0} + e^{\ell_1} + \cdots
      + e^{\ell_{V-1}}\bigr)
      - e^{\ell_1}\,e^{\ell_2}}
    {\bigl(e^{\ell_0} + e^{\ell_1} + \cdots + e^{\ell_{V-1}}\bigr)^{2}}
    \notag                                                    \\
     & = -\,\frac{e^{\ell_1}}{S}\cdot\frac{e^{\ell_2}}{S}
    \;=\; -\,p_1\,p_2.
    \label{eq:q12}
  \end{align}

  \paragraph{Entries $\pdv{p_2}{\ell_1}$ and $\pdv{p_2}{\ell_2}$}
  We perform the same two computations with the indices swapped. For $\pdv{p_2}{\ell_1}$, $u = e^{\ell_2}$ is constant with respect to $\ell_1$, and $\pdv{v}{\ell_1} = e^{\ell_1}$ by \eqref{eq:q11-v}. For $\pdv{p_2}{\ell_2}$, both the numerator and the denominator vary, matching \eqref{eq:q11}:
  \begin{align}
    \pdv{p_2}{\ell_1}
     & = \frac{0 \cdot S - e^{\ell_2}\,e^{\ell_1}}{S^{2}}
    = -\,p_2\,p_1,
    \label{eq:q21}                                            \\
    \pdv{p_2}{\ell_2}
     & = \frac{e^{\ell_2}\,S - e^{\ell_2}\,e^{\ell_2}}{S^{2}}
    = p_2\,(1 - p_2).
    \label{eq:q22}
  \end{align}

  \paragraph{The pattern}
  Arranging these four entries as a block of the Jacobian yields:
  \begin{equation}
    \begin{pmatrix}
      \pdv{p_1}{\ell_1} & \pdv{p_1}{\ell_2} \\[6pt]
      \pdv{p_2}{\ell_1} & \pdv{p_2}{\ell_2}
    \end{pmatrix}
    =
    \begin{pmatrix}
      p_1\,(1-p_1) & -\,p_1\,p_2  \\[2pt]
      -\,p_2\,p_1  & p_2\,(1-p_2)
    \end{pmatrix}:
    \label{eq:softmax-block}
  \end{equation}
  this block contains diagonal entries $p_i\,(1 - p_i)$ and off-diagonal entries $-\,p_i\,p_j$. The derivation of the general entry follows the same quotient rule. Each derivative of $v$ above illustrates the general fact that exactly one term of the expanded sum $S$ depends on $\ell_j$. In delta form, we differentiate each term using \eqref{eq:entry-dv} and collapse the sum via sifting \eqref{eq:sift}:
  \begin{equation}
    \pdv{S}{\ell_j}
    = \sum_{a<V} \pdv{}{\ell_j}\,e^{\ell_a}
    = \sum_{a<V} e^{\ell_a}\,\kron{a}{j}
    = e^{\ell_j}.
    \label{eq:softmax-dS}
  \end{equation}
  The diagonal case $j = i$ repeats \eqref{eq:q11} because the numerator also varies:
  \begin{align}
    \pdv{p_i}{\ell_i}
     & = \frac{e^{\ell_i}\,S - e^{\ell_i}\,e^{\ell_i}}{S^2}
    = \frac{e^{\ell_i}}{S}
    - \left(\frac{e^{\ell_i}}{S}\right)^{\!2} \notag        \\
     & = p_i - p_i^2 = p_i\,(1 - p_i).
    \label{eq:softmax-diag}
  \end{align}
  The off-diagonal case $j \neq i$ repeats \eqref{eq:q12} because the numerator $e^{\ell_i}$ is constant with respect to $\ell_j$:
  \begin{equation}
    \pdv{p_i}{\ell_j}
    = \frac{0 \cdot S - e^{\ell_i}\,e^{\ell_j}}{S^2}
    = -\,p_i\,p_j.
    \label{eq:softmax-offdiag}
  \end{equation}
\end{derivation}

\begin{result}
  The two cases combine into one expression via the Kronecker delta
  \eqref{eq:kron}, which selects the diagonal case:
  \begin{equation}
    \pdv{p_i}{\ell_j} \;=\; p_i\,\bigl(\kron{i}{j} - p_j\bigr).
    \label{eq:softmax-jac}
  \end{equation}
  Check against \eqref{eq:softmax-diag} and \eqref{eq:softmax-offdiag}:
  $j = i$ gives $p_i\,(1 - p_i)$, and $j \neq i$ gives $-\,p_i\,p_j$.
\end{result}

\subsection{Backward w.r.t.\ the logits}

\begin{derivation}
  We apply the upstream gradient $\dL{p}$ through the chain rule \eqref{eq:chain} summed over the row, using one column of the Jacobian \eqref{eq:softmax-jac} for each output $i$. The summation index $j$ runs over the upstream coordinates, while the free index $i$ denotes the logit gradient entry. Expanding this sum yields:
  \begin{align}
    \dL{\ell_i}
     & = \sum_{j<V} \dL{p_j}\,\pdv{p_j}{\ell_i} \notag      \\
     & = \dL{p_0}\pdv{p_0}{\ell_i}
    + \dL{p_1}\pdv{p_1}{\ell_i}
    + \dL{p_2}\pdv{p_2}{\ell_i} \notag                      \\
     & \qquad + \cdots + \dL{p_{V-1}}\pdv{p_{V-1}}{\ell_i}.
    \label{eq:softmax-chain}
  \end{align}
  We substitute \eqref{eq:softmax-diag} and \eqref{eq:softmax-offdiag}, separating the diagonal term $j = i$ from the off-diagonal terms:
  \begin{align}
    \dL{\ell_i}
     & = \dL{p_i}\,p_i\,(1 - p_i)
    + \sum_{j \neq i} \dL{p_j}\,(-\,p_j\,p_i) \notag   \\
     & = p_i\,\dL{p_i}
    - p_i\,p_i\,\dL{p_i}
    - p_i \sum_{j \neq i} \dL{p_j}\,p_j.
    \intertext{Because the middle term corresponds exactly to the missing $j = i$ term of the sum, we absorb it back into the summation:}
    \dL{\ell_i}
     & = p_i\,\dL{p_i} - p_i \sum_{j<V} \dL{p_j}\,p_j,
  \end{align}
  where the remaining sum is identical for every $i$ in the row. Expanding this term yields:
  \begin{multline}
    \sum_{j<V} \dL{p_j}\,p_j
    = \dL{p_0}\,p_0 + \dL{p_1}\,p_1 \\
    + \dL{p_2}\,p_2 + \cdots + \dL{p_{V-1}}\,p_{V-1}.
    \label{eq:softmax-dot}
  \end{multline}
\end{derivation}

\begin{result}
  \begin{equation}
    \dL{\ell_i} \;=\;
    p_i \left(\dL{p_i} - \sum_{j<V} \dL{p_j}\,p_j\right),
    \qquad i < V,
    \label{eq:softmax-bwd}
  \end{equation}
  and $\dL{\ell_i} = 0$ for the padded region $i \geq V$. When we sum \eqref{eq:softmax-bwd} over the row, the leading factor $p_i$ distributes into both terms. The probabilities multiplying the constant dot product sum to one, leaving two copies of the dot product:
  \begin{align}
    \sum_{i<V} \dL{\ell_i}
     & = \sum_{i<V} p_i\,\dL{p_i}
    - \Bigl(\sum_{i<V} p_i\Bigr)
    \sum_{j<V} \dL{p_j}\,p_j \notag \\
     & = \sum_{i<V} p_i\,\dL{p_i}
    - \sum_{j<V} \dL{p_j}\,p_j
    \;=\; 0.
    \label{eq:softmax-zerosum}
  \end{align}
  The gradient row sums to zero, which reflects the shift invariance \eqref{eq:softmax-shift}.
\end{result}

\subsection{Fused softmax + cross-entropy}

This composition represents the primary use case for the model. Backpropagating through \eqref{eq:ce-fwd} and \eqref{eq:softmax-fwd} simultaneously collapses the Jacobian.

\begin{derivation}
  We write the chain rule per logit coordinate, restoring the full indices. The summation index $i$ runs over the probabilities, while the free index $j$ denotes the logit gradient entry:
  \begin{equation}
    \dL{\ell_{bt,j}}
    \;=\;
    \sum_{i<V} \dL{p_{bt,i}}\,\pdv{p_{bt,i}}{\ell_{bt,j}}.
    \label{eq:fused-chain}
  \end{equation}
  We substitute the cross-entropy gradient \eqref{eq:ce-bwd} for the first factor and the Jacobian \eqref{eq:softmax-jac} for the second factor into the combined equation:
  \begin{align}
    \dL{\ell_{bt,j}}
     & = \sum_{i<V}
    \left(-\frac{\ind{i = y_{bt}}}{p_{bt,i}}\right)
    p_{bt,i}\,\bigl(\kron{i}{j} - p_{bt,j}\bigr)
    \intertext{The factor $p_{bt,i}$ in the denominator cancels with the factor $p_{bt,i}$ from the Jacobian:}
     & = -\sum_{i<V} \ind{i = y_{bt}}\,
    \bigl(\kron{i}{j} - p_{bt,j}\bigr)
    \intertext{and the one-hot indicator isolates the single term where $i = y_{bt}$:}
     & = -\bigl(\kron{y_{bt}}{j} - p_{bt,j}\bigr)
    = p_{bt,j} - \ind{j = y_{bt}}.
    \label{eq:fused-collapse}
  \end{align}
\end{derivation}

\begin{result}
  \begin{equation}
    \dL{\ell_{bt,j}} \;=\; p_{bt,j} - \ind{j = y_{bt}},
    \qquad j < V,
    \label{eq:fused-bwd}
  \end{equation}
  and $\dL{\ell_{bt,j}} = 0$ for the padded region $j \geq V$. The Jacobian collapses entirely, removing the sum over the row in \eqref{eq:fused-bwd}.

  We verify this result via the general backward formula \eqref{eq:softmax-bwd}. With the cross-entropy upstream gradient \eqref{eq:ce-bwd}, the dot product evaluates to:
  \begin{equation}
    \sum_{j<V} \dL{p_{bt,j}}\,p_{bt,j}
    = \sum_{j<V} \left(-\frac{\ind{j = y_{bt}}}{p_{bt,j}}\right) p_{bt,j}
    = -1,
    \label{eq:fused-dot}
  \end{equation}
  which is a constant. For a one-hot target, we determine the row-wide weighted sum before reading any element. Substituting $-1$ for the dot product in \eqref{eq:softmax-bwd}, the factor $p_{bt,i}$ outside the parentheses cancels the reciprocal inside:
  \begin{align}
    \dL{\ell_{bt,i}}
     & = p_{bt,i}\left(\dL{p_{bt,i}} - (-1)\right) \notag      \\
     & = p_{bt,i}
    \left(-\frac{\ind{i = y_{bt}}}{p_{bt,i}} + 1\right) \notag \\
     & = p_{bt,i} - \ind{i = y_{bt}},
    \label{eq:fused-check}
  \end{align}
  demonstrating that chaining through the Jacobian directly and substituting into the general backward formula represent two equivalent routes to the same gradient. The zero-sum property of \S\ref{sec:softmax} also holds because the probabilities sum to one, and the indicator fires exactly once per row:
  \begin{equation}
    \sum_{j<V} \bigl(p_{bt,j} - \ind{j = y_{bt}}\bigr)
    \;=\; 1 - 1 \;=\; 0.
    \label{eq:fused-zerosum}
  \end{equation}

  The collapse does not require a one-hot target. It requires only the standard-sum gradient \eqref{eq:ce-std-bwd} and the assumption that the target is a probability distribution satisfying $\sum_{i<V} y_{bt,i} = 1$. Repeating \eqref{eq:fused-chain} with a general target distribution:
  \begin{align}
    \dL{\ell_{bt,j}}
     & = \sum_{i<V}
    \left(-\frac{y_{bt,i}}{p_{bt,i}}\right)
    p_{bt,i}\,\bigl(\kron{i}{j} - p_{bt,j}\bigr) \notag \\
     & = -\,y_{bt,j} + p_{bt,j}\sum_{i<V} y_{bt,i}
    \;=\; p_{bt,j} - y_{bt,j},
    \label{eq:fused-general}
  \end{align}
  where the one-hot result \eqref{eq:fused-bwd} is the special case where $y_{bt,j} = \ind{j = y_{bt}}$. Thus, softmax paired with cross-entropy always backpropagates as \emph{predicted minus target}.
\end{result}

% =============================================================================
\section{Linear Layer (Matmul)}\label{sec:matmul}

We express the forward pass as $Y = XW^{\top} + b$ with $X \in \R^{(B\cdot T)\times C_{\text{in}}}$, $W \in \R^{C_{\text{out}}\times C_{\text{in}}}$, and $b \in \R^{C_{\text{out}}}$.

\begin{derivation}
  We write the equations in index form: $n$ runs over the $N = B \cdot T$ flattened rows ($n = bt$), $j$ runs over $C_{\text{in}}$, and $i$ runs over $C_{\text{out}}$. We interpret every matrix entry as (row, column): $X_{nj}$ is (position $n$, in-channel $j$); $W_{ij}$ is (out-channel $i$, in-channel $j$); $Y_{ni}$ is (position $n$, out-channel $i$); and $b_i$ has the single index (out-channel $i$).

  Differentiation requires a second, independent set of index names. To determine how $Y_{ni}$ changes when one entry of $X$ varies, we must allow the perturbed entry to be any element. Thus, the probe entry receives fresh indices: $X_{mk}$ (position $m$, in-channel $k$), and similarly $W_{kl}$ (out-channel $k$, in-channel $l$) and $b_k$. Keeping the target indices $(n, i)$ and the probe indices distinct evaluates every combination of output and input simultaneously.

  \paragraph{The transpose under differentiation}
  By the transpose definition \eqref{eq:transpose}, we write $(W^{\top})_{ji} = W_{ij}$, which relabels the same independent variables. Applying \eqref{eq:entry-dv} through the relabeled indices, we find that the transpose changes only which index of $W$ we compare with the probe index:
  \begin{equation}
    \pdv{(W^{\top})_{ji}}{W_{kl}}
    \;=\;
    \pdv{W_{ij}}{W_{kl}}
    \;=\;
    \kron{i}{k}\,\kron{j}{l}.
    \label{eq:transpose-dv}
  \end{equation}
  After absorbing \eqref{eq:transpose} and writing out the broadcast bias \eqref{eq:broadcast}, we express the forward pass in index form as:
  \begin{align}
    Y_{ni}
     & = \sum_{j<C_{\text{in}}} X_{nj}\,(W^{\top})_{ji} + b_i
    = \sum_{j<C_{\text{in}}} X_{nj}\,W_{ij} + b_i \notag      \\
     & = X_{n0}\,W_{i0} + X_{n1}\,W_{i1}
    + \cdots + X_{n,C_{\text{in}}-1}\,W_{i,C_{\text{in}}-1}
    + b_i.
    \label{eq:matmul-fwd-idx}
  \end{align}

  \paragraph{Local derivative $\pdv{Y}{X}$}
  With target $Y_{ni}$ (position $n$, out-channel $i$) and probe $X_{mk}$ (position $m$, in-channel $k$), we consider four free indices, representing every pair of output and input entries. We differentiate the expanded form \eqref{eq:matmul-fwd-idx} term by term. Because $W$ and $b$ act as constants, and each $X_{nj}$ obeys \eqref{eq:entry-dv}, we obtain:
  \begin{align}
    \pdv{Y_{ni}}{X_{mk}}
     & = \sum_{j<C_{\text{in}}} W_{ij}\,\pdv{X_{nj}}{X_{mk}}
    = \sum_{j<C_{\text{in}}} W_{ij}\,\kron{n}{m}\,\kron{j}{k} \notag \\
     & = \kron{n}{m} \sum_{j<C_{\text{in}}} W_{ij}\,\kron{j}{k}
    = \kron{n}{m}\,W_{ik}.
    \label{eq:matmul-dydx}
  \end{align}
  Because the row delta $\kron{n}{m}$ does not depend on $j$, we factor it out of the sum. The column delta sifts the sum to $j = k$ via \eqref{eq:sift}. The delta $\kron{n}{m}$ survives because we do not sum over $n$ or $m$ here. This term remains for the chain rule, indicating that row $n$ of $Y$ depends only on row $m = n$ of $X$.

  \paragraph{Local derivative $\pdv{Y}{W}$}
  With target $Y_{ni}$ and probe $W_{kl}$ (out-channel $k$, in-channel $l$), $X$ acts as the constant and each $W_{ij}$ obeys \eqref{eq:transpose-dv}:
  \begin{align}
    \pdv{Y_{ni}}{W_{kl}}
     & = \sum_{j<C_{\text{in}}} X_{nj}\,\pdv{W_{ij}}{W_{kl}}
    = \sum_{j<C_{\text{in}}} X_{nj}\,\kron{i}{k}\,\kron{j}{l} \notag \\
     & = \kron{i}{k} \sum_{j<C_{\text{in}}} X_{nj}\,\kron{j}{l}
    = \kron{i}{k}\,X_{nl}.
    \label{eq:matmul-dydw}
  \end{align}
  This calculation mirrors the previous derivation with the roles swapped: the out-channel delta $\kron{i}{k}$ does not depend on $j$ and factors out, while the in-channel delta sifts the sum over $j$ to $l$.

  \paragraph{Local derivative $\pdv{Y}{b}$ and its broadcast behavior}
  The bias term in \eqref{eq:matmul-fwd-idx} is $b_i$. Because $b$ is a vector, \eqref{eq:entry-dv} requires only one index pair:
  \begin{equation}
    \pdv{Y_{ni}}{b_k} \;=\; \pdv{b_i}{b_k} \;=\; \kron{i}{k},
    \label{eq:matmul-dydb}
  \end{equation}
  which does not depend on $n$. This independence reflects the broadcast operation: the same $b_k$ appears in every one of the $N$ rows of $Y$, meaning that $b_k$ affects $N$ distinct output coordinates. While this local derivative resembles the scalar example (\S\ref{sec:example}, $\pdv{z}{b} = 1$), the broadcast behavior introduces a sum in the chain rule that we must not omit.

  \paragraph{Chain rule}
  Because $L$ depends on every entry of $Y$, the chain rule \eqref{eq:chain} sums over both output indices $(n, i)$. The probe indices act as the free indices, specifying the gradient entry we compute:
  \begin{equation}
    \dL{\theta}
    \;=\;
    \sum_{n<N} \sum_{i<C_{\text{out}}}
    \dL{Y_{ni}}\,\pdv{Y_{ni}}{\theta}.
    \label{eq:matmul-chain}
  \end{equation}
  For the input gradient, we substitute \eqref{eq:matmul-dydx} into the chain rule. The surviving row delta sifts the sum over $n$ to $m$. Because no delta constrains $i$, the sum over $i$ remains as a contraction with $W$:
  \begin{align}
    \dL{X_{mk}}
     & = \sum_{n<N} \sum_{i<C_{\text{out}}}
    \dL{Y_{ni}}\,\kron{n}{m}\,W_{ik} \notag          \\
     & = \sum_{i<C_{\text{out}}} \dL{Y_{mi}}\,W_{ik}
    = \Bigl(\dL{Y}\,W\Bigr)_{mk}.
    \label{eq:matmul-dx-idx}
  \end{align}
  This expression has free indices $(m, k)$ on both sides, corresponding to row $m$, in-channel $k$ of the input gradient. The final step applies the recognition rule of \eqref{eq:matprod}: the summed index $i$ is the column of $\dL{Y}$ (in $\dL{Y_{mi}}$) and the row of $W$ (in $W_{ik}$), meaning the sum represents a matrix product without transposing either factor.

  For the weight gradient, we substitute \eqref{eq:matmul-dydw} into the chain rule. The out-channel delta sifts the sum over $i$ to $k$, but because no delta constrains $n$, the sum over $n$ survives:
  \begin{align}
    \dL{W_{kl}}
     & = \sum_{n<N} \sum_{i<C_{\text{out}}}
    \dL{Y_{ni}}\,\kron{i}{k}\,X_{nl} \notag \\
     & = \sum_{n<N} \dL{Y_{nk}}\,X_{nl}
    = \left(\dL{Y}^{\top} X\right)_{kl}.
    \label{eq:matmul-dw-idx}
  \end{align}
  This expression has free indices $(k, l)$, which match the out-channel $k$ and in-channel $l$ layout of $W$. Here, the recognition rule of \eqref{eq:matprod} requires a transpose because the summed index $n$ is the row of both $\dL{Y}$ and $X$. Transposing the left factor yields $(\dL{Y}^{\top})_{kn} = \dL{Y_{nk}}$ via \eqref{eq:transpose}, which expresses the sum as a matrix product.

  For the bias gradient, we substitute \eqref{eq:matmul-dydb} into the chain rule. The delta sifts the sum over $i$. Because the local derivative does not depend on $n$, the sum over $n$ survives. Expanding this sum yields:
  \begin{align}
    \dL{b_k}
     & = \sum_{n<N} \sum_{i<C_{\text{out}}}
    \dL{Y_{ni}}\,\kron{i}{k}
    = \sum_{n<N} \dL{Y_{nk}} \notag              \\
     & = \dL{Y_{0k}} + \dL{Y_{1k}} + \dL{Y_{2k}}
    + \cdots + \dL{Y_{N-1,k}}.
    \label{eq:matmul-db-idx}
  \end{align}
  This demonstrates the consequence of the broadcast behavior described in \eqref{eq:matmul-dydb}: although the local derivative is simply $\kron{i}{k}$, the gradient requires a summation over all $B \cdot T$ rows. In the scalar example (\S\ref{sec:example}), the same relation holds with $N = 1$. The sum still exists there, but collapses to a single term, yielding $\dL{b} = \dL{z}$.
\end{derivation}

\begin{result}
  \begin{align}
    \dL{X}   & =\; \dL{Y}\,W,              \\
    \dL{W}   & =\; \dL{Y}^{\top} X,        \\
    \dL{b_k} & =\; \sum_{n<N} \dL{Y_{nk}}.
    \label{eq:matmul-bwd}
  \end{align}
  We verify the dimensions of each gradient with $\dL{Y} \in \R^{N \times C_{\text{out}}}$:
  $(N \times C_{\text{out}})(C_{\text{out}} \times C_{\text{in}})
    = N \times C_{\text{in}}$ for $\dL{X}$;
  $(C_{\text{out}} \times N)(N \times C_{\text{in}})
    = C_{\text{out}} \times C_{\text{in}}$ for $\dL{W}$;
  $\dL{b} \in \R^{C_{\text{out}}}$.

  The reduction over the $B \cdot T$ dimension occurs in $\dL{W}$ and $\dL{b}$ because every position contributes to the shared parameters (using the $\mathrel{+}=$ convention of \S\ref{sec:notation}). In contrast, we compute a separate gradient row for each input row in $\dL{X}$.

  Because the forward pass contains the transpose ($Y = XW^{\top}$), two consequences follow. First, computing the input gradient requires no transpose ($\dL{X} = \dL{Y}\,W$, rather than $\dL{Y}\,W^{\top}$). Second, we obtain $\dL{W}$ directly in the original layout of $W$ ($C_{\text{out}} \times C_{\text{in}}$). Both properties derive from \eqref{eq:transpose-dv} rather than arbitrary conventions.
\end{result}

% =============================================================================
\section{GELU}\label{sec:gelu}

The Gaussian Error Linear Unit (GELU) activation function, proposed by Hendrycks and Gimpel \cite{gelu}, scales the input by the cumulative normal distribution function $\gelu(x) = x \Phi(x)$. Because the exact CDF involves the non-elementary error function, we express the forward pass using the fast hyperbolic tangent approximation introduced by Page \cite{page} and adopted by OpenAI in GPT-2 and GPT-3:
\begin{equation}
  \gelu(x) \;=\; \tfrac{1}{2}\,x\left(1 + \tanh\!\left[\sqrt{2/\pi}\,
      (x + 0.044715\,x^{3})\right]\right).
  \label{eq:gelu-fwd}
\end{equation}

We define the inner argument $u(x)$ to simplify the chain rule:
\begin{equation}
  \begin{aligned}
    u(x)     & = \sqrt{2/\pi}\,\bigl(x + 0.044715\,x^{3}\bigr), \\
    \gelu(x) & = \tfrac{1}{2}\,x\,\bigl(1 + \tanh u\bigr).
  \end{aligned}
  \label{eq:gelu-u}
\end{equation}

\begin{derivation}
  We compute the scalar derivative $\gelu'(x)$. The result describes its role in the chain rule.

  \paragraph{Product rule}
  We apply the product rule to \eqref{eq:gelu-u}:
  \begin{equation}
    \gelu'(x)
    \;=\;
    \tfrac{1}{2}\bigl(1 + \tanh u\bigr)
    + \tfrac{1}{2}\,x\,\pdv{}{x}\tanh u.
    \label{eq:gelu-product}
  \end{equation}

  \paragraph{The tanh derivative via $\operatorname{sech}^2$}
  The derivative of $\tanh$ is $\operatorname{sech}^{2}$, which requires chaining the derivative of the inner function:
  \begin{equation}
    \pdv{}{x}\tanh u
    \;=\;
    \operatorname{sech}^{2}(u)\,u'(x),
    \label{eq:gelu-sech}
  \end{equation}
  with the polynomial derivative of the inner argument:
  \begin{align}
    u'(x)
     & = \sqrt{2/\pi}\,\bigl(1 + 3 \cdot 0.044715\,x^{2}\bigr) \notag \\
     & = \sqrt{2/\pi}\,\bigl(1 + 0.134145\,x^{2}\bigr).
    \label{eq:gelu-du}
  \end{align}

  \paragraph{Converting $\operatorname{sech}^2$ to $1 - \tanh^2$}
  Dividing the hyperbolic Pythagorean identity $\cosh^{2} u - \sinh^{2} u = 1$ by $\cosh^{2} u$ yields:
  \begin{equation}
    1 - \tanh^{2} u \;=\; \operatorname{sech}^{2} u,
    \label{eq:sech-tanh}
  \end{equation}
  so \eqref{eq:gelu-sech} needs no separate $\operatorname{sech}$ evaluation:
  \begin{equation}
    \pdv{}{x}\tanh u
    \;=\;
    \bigl(1 - \tanh^{2} u\bigr)\,u'(x).
    \label{eq:gelu-tanhprime}
  \end{equation}

  \paragraph{Simplifying the derivative}
  Because $\tanh$ differentiates into a function of itself ($\tanh' = 1 - \tanh^{2}$), and $\tanh u$ appears in the forward expression \eqref{eq:gelu-u}, we solve for it when $x \neq 0$:
  \begin{equation}
    1 + \tanh u = \frac{2\,\gelu(x)}{x}
    \;\Longrightarrow\;
    \tanh u = \frac{2\,\gelu(x)}{x} - 1.
    \label{eq:gelu-self}
  \end{equation}
  We substitute \eqref{eq:gelu-tanhprime} into \eqref{eq:gelu-product}. This ensures that every quantity in the derivative is either the input $x$ or the value $\tanh u$ computed during the forward pass:
  \begin{align}
    \gelu'(x)
     & = \tfrac{1}{2}\bigl(1 + \tanh u\bigr)
    + \tfrac{1}{2}\,x\,\bigl(1 - \tanh^{2} u\bigr)\,u'(x) \notag \\
     & = \frac{\gelu(x)}{x}
    + \tfrac{1}{2}\,x\,\bigl(1 - \tanh^{2} u\bigr)\,u'(x).
    \label{eq:gelu-rolled}
  \end{align}
\end{derivation}

\begin{result}
  \begin{equation}
    \gelu'(x)
    \;=\;
    \tfrac{1}{2}\bigl(1 + \tanh u\bigr)
    + \tfrac{1}{2}\,x\,\bigl(1 - \tanh^{2} u\bigr)\,u'(x),
    \label{eq:gelu-bwd}
  \end{equation}
  where we compute $u$ and $u'$ via \eqref{eq:gelu-u} and \eqref{eq:gelu-du}. A single evaluation of $\tanh u$ serves both terms because the self-reference $\tanh' = 1 - \tanh^{2}$ \eqref{eq:sech-tanh} replaces $\operatorname{sech}$, and \eqref{eq:gelu-self} expresses the first term as the forward output divided by $x$. At $x = 0$, we use \eqref{eq:gelu-bwd} directly: $u(0) = 0$ and $\tanh 0 = 0$ yield $\gelu'(0) = \tfrac{1}{2}$, avoiding the singularity in the $\gelu(x)/x$ term of \eqref{eq:gelu-rolled}.

  GELU operates elementwise: each output coordinate depends only on its corresponding input coordinate. By \eqref{eq:entry-dv}, the Jacobian is diagonal:
  \begin{equation}
    \pdv{\gelu(x_a)}{x_b} \;=\; \gelu'(x_a)\,\kron{a}{b},
    \label{eq:gelu-diag}
  \end{equation}
  and the chain rule \eqref{eq:chain} collapses the sum via sifting \eqref{eq:sift} to a pointwise product: $\dL{x_a} = \dL{y_a}\,\gelu'(x_a)$.
\end{result}

% =============================================================================
\section{LayerNorm}\label{sec:layernorm}

Layer normalization, proposed by Ba et al.\ \cite{layernorm}, stabilizes training dynamics by normalizing activations across the channel dimension within each sequence position. We compute the forward pass at each position $x \in \R^{C}$ with scale $\gamma$ and shift $\beta$ as:
\begin{equation}
  \begin{aligned}
    \mu       & = \tfrac{1}{C}\sum_j x_j,
    \qquad
    \sigma^2 = \tfrac{1}{C}\sum_j (x_j - \mu)^2 + \varepsilon, \\
    \hat{x}_i & = \frac{x_i - \mu}{\sqrt{\sigma^2}},           \\
    y_i       & = \gamma_i\,\hat{x}_i + \beta_i.
  \end{aligned}
  \label{eq:ln-fwd}
\end{equation}

\begin{derivation}
  We compute the partial derivatives of the output $y_i$ with respect to its parameters and inputs. The output $y_i$ depends on every $x_j$ in the row through the statistics $\mu$ and $\sigma^2$.

  \paragraph{Parameter partials}
  The output $y_i$ depends only on the scale and shift parameters of its corresponding channel:
  \begin{equation}
    \pdv{y_i}{\gamma_i} = \frac{x_i - \mu}{\sqrt{\sigma^2}},
    \qquad
    \pdv{y_i}{\beta_i} = 1.
    \label{eq:ln-dydparam}
  \end{equation}

  \paragraph{Input partials}
  To find the Jacobian $\pdv{y_j}{x_i}$, we first differentiate the row statistics. Each input element $x_i$ contributes $1/C$ to the mean and $2(x_i - \mu)/C$ to the variance:
  \begin{equation}
    \pdv{\mu}{x_i} = \frac{1}{C},
    \qquad
    \pdv{\sigma^2}{x_i} = \frac{2(x_i - \mu)}{C}.
    \label{eq:ln-dstats}
  \end{equation}
  We differentiate the normalized value $\hat{x}_j$ with respect to $x_i$ using the quotient rule on $\hat{x}_j = (x_j - \mu)(\sigma^2)^{-1/2}$:
  \begin{align}
    \pdv{\hat{x}_j}{x_i}
     & = \pdv{}{x_i} \left( \frac{x_j - \mu}{\sqrt{\sigma^2}} \right) \notag                                                             \\
     & = \frac{(\kron{j}{i} - \frac{1}{C})\sqrt{\sigma^2} - (x_j - \mu)\frac{1}{2}(\sigma^2)^{-1/2}\pdv{\sigma^2}{x_i}}{\sigma^2} \notag \\
     & = \frac{1}{\sqrt{\sigma^2}} \left( \kron{j}{i} - \frac{1}{C} - \frac{(x_j - \mu)(x_i - \mu)}{C\sigma^2} \right).
    \label{eq:ln-dxhatdx}
  \end{align}
  The partial derivative $\pdv{y_j}{x_i} = \gamma_j \pdv{\hat{x}_j}{x_i}$ describes how a change in $x_i$ affects every output in the row.
\end{derivation}

\begin{result}
  We compute the total gradient for each quantity by summing its contributions to the loss across all outputs $y_j$, substituting the partial derivatives derived above:
  \begin{align}
    \dL{\gamma_i} & = \dL{y_i}\,\pdv{y_i}{\gamma_i} = \dL{y_i}\,\frac{x_i - \mu}{\sqrt{\sigma^2}}, \\
    \dL{\beta_i}  & = \dL{y_i}\,\pdv{y_i}{\beta_i} = \dL{y_i}.
  \end{align}
  For the input gradient $\dL{x_i}$, we sum over the entire row $j \in \{0, \dots, C-1\}$. We expand the Kronecker delta $\kron{j}{i}$ into its indicator form $\ind{j=i}$ to clarify the sifting process:
  \begin{align}
    \dL{x_i}
     & = \sum_j \dL{y_j} \pdv{y_j}{x_i} \notag                                                                                                            \\
     & = \sum_j \dL{y_j} \left[ \frac{\gamma_j}{\sqrt{\sigma^2}} \left( \kron{j}{i} - \frac{1}{C} - \frac{\hat{x}_i\,\hat{x}_j}{C} \right) \right] \notag \\
     & = \frac{1}{\sqrt{\sigma^2}} \left( \sum_j \gamma_j\,\dL{y_j}\,\ind{j=i} - \frac{1}{C} \sum_j \gamma_j\,\dL{y_j} \right. \notag                     \\
     & \qquad\qquad \left. - \frac{\hat{x}_i}{C} \sum_j \gamma_j\,\dL{y_j}\,\hat{x}_j \right) \notag                                                      \\
     & = \frac{1}{\sqrt{\sigma^2}} \left( \gamma_i\,\dL{y_i} - \frac{1}{C} \sum_j \gamma_j\,\dL{y_j} \right. \notag                                       \\
     & \qquad\qquad \left. - \frac{\hat{x}_i}{C} \sum_j \gamma_j\,\dL{y_j}\,\hat{x}_j \right).
    \label{eq:ln-bwd}
  \end{align}
  The row-wide sums $\sum \gamma_j \dL{y_j} / C$ and $\sum \gamma_j \dL{y_j} \hat{x}_j / C$ match the row statistics $s_1$ and $s_2$ that the first pass of \texttt{layernorm\_bwd} computes.
\end{result}

% =============================================================================
\section{Attention}\label{sec:attention}

Multi-head self-attention, the central building block of the Transformer architecture \cite{attention}, allows sequence elements to dynamically route information by computing scaled dot-product attention over queries, keys, and values. We express the forward pass for each head as $A = \softmax\!\big(QK^{\top}/\sqrt{h} + M\big)$ and $Y = AV$, with causal mask $M$ and projections $Q = XW_Q$, $K = XW_K$, and $V = XW_V$.

\begin{derivation}
  We write the equations in index form: $i$ runs over query positions $T$ ($i < T$), $j$ runs over key/value positions $T$ ($j < T$), and $d$ runs over the head dimension $h$ ($d < h$). We drop the batch and head indices throughout this section because each head is independent, allowing us to derive the gradients per-head. The channel size of the query and key projections within each head is the head dimension $h$, corresponding to the standard notation $d_k$ in the literature (which scales the attention scores by $1/\sqrt{d_k}$). We obtain the head dimension as $h = C/H$, where $C$ is the total channel width (embedding dimension) and $H$ is the number of attention heads (see Table~\ref{tab:notation} in \S\ref{sec:notation}). We write the scaling factor as $1/\sqrt{h}$ to match the head dimension symbol in the kernel code. We interpret each entry as (row, column): $Q_{id}$ is (query position $i$, channel $d$); $K_{jd}$ is (key position $j$, channel $d$); $V_{jd}$ is (value position $j$, channel $d$); $Y_{id}$ is (output position $i$, channel $d$); $S_{ij}$ is the raw score at (query position $i$, key position $j$); and $A_{ij}$ is the attention probability at (query position $i$, key position $j$).

  The forward pass computes the raw causal score matrix $S \in \R^{T \times T}$ as:
  \begin{equation}
    S_{ij} \;=\; \begin{cases}
      \frac{1}{\sqrt{h}} \sum_{d<h} Q_{id} K_{jd} & j \leq i, \\
      -\infty                                     & j > i.
    \end{cases}
    \label{eq:att-score-fwd}
  \end{equation}
  We obtain the attention probability matrix $A \in \R^{T \times T}$ by applying the row-wise softmax to $S$:
  \begin{equation}
    A_{ij} \;=\; \softmax(S_{i,\cdot})_j \;=\; \frac{e^{S_{ij} - L_i}}{\sum_{l \leq i} e^{S_{il} - L_i}},
    \label{eq:att-prob-fwd}
  \end{equation}
  where $L_i = \log \sum_{l \leq i} e^{S_{il}}$ is the log-sum-exp (LSE) value for row $i$. By definition, $A_{ij} = 0$ for $j > i$. The head output $Y \in \R^{T \times h}$ is the matrix product:
  \begin{equation}
    Y_{id} \;=\; \sum_{j<T} A_{ij} V_{jd} \quad \implies \quad Y \;=\; A V.
    \label{eq:att-output-fwd}
  \end{equation}

  Differentiation requires a second, independent set of index names to avoid collisions. We choose the probe indices $m$ for position and $k$ for channel.

  \paragraph{Local derivative $\pdv{Y}{V}$}
  With target $Y_{id}$ (query position $i$, channel $d$) and probe $V_{mk}$ (key position $m$, channel $k$), we differentiate:
  \begin{align}
    \pdv{Y_{id}}{V_{mk}}
     & = \pdv{}{V_{mk}} \sum_{a<T} A_{ia}\,V_{ad} \notag    \\
     & = \sum_{a<T} A_{ia}\,\kron{a}{m}\,\kron{d}{k} \notag \\
     & = A_{im}\,\kron{d}{k}.
    \label{eq:att-dydv}
  \end{align}
  The Kronecker deltas act as coordinate checks: $\kron{d}{k}$ restricts the channel index to $d=k$, while $\kron{a}{m}$ collapses the sum over position $a$ to the single term where $a=m$. We chain the upstream gradient $\dL{Y}$ using the multivariable chain rule \eqref{eq:chain}:
  \begin{align}
    \dL{V_{mk}}
     & = \sum_{i<T} \sum_{d<h} \dL{Y_{id}}\,\pdv{Y_{id}}{V_{mk}} \notag \\
     & = \sum_{i<T} \sum_{d<h} \dL{Y_{id}}\,A_{im}\,\kron{d}{k} \notag  \\
     & = \sum_{i<T} A_{im}\,\dL{Y_{ik}} \notag                          \\
     & = \sum_{i<T} (A^{\top})_{mi}\,\dL{Y_{ik}}
    = \Bigl( A^{\top}\,\dL{Y} \Bigr)_{mk}.
    \label{eq:att-chain-dv}
  \end{align}
  In the third step, the channel delta $\kron{d}{k}$ sifts the sum over $d$ to $k$. In the fourth step, we write $A_{im} = (A^{\top})_{mi}$ to transpose the attention matrix; this aligns the summed index $i$ as the inner dimension of the matrix product, matching the recognition rule of \eqref{eq:matprod} to yield $\dL{V} = A^{\top}\,\dL{Y}$.

  \paragraph{Local derivative $\pdv{Y}{A}$}
  With target $Y_{id}$ and probe $A_{mj}$ (query position $m$, key position $j$):
  \begin{align}
    \pdv{Y_{id}}{A_{mj}}
     & = \pdv{}{A_{mj}} \sum_{a<T} A_{ia}\,V_{ad} \notag    \\
     & = \sum_{a<T} \kron{i}{m}\,\kron{a}{j}\,V_{ad} \notag \\
     & = \kron{i}{m}\,V_{jd}.
    \label{eq:att-dyda}
  \end{align}
  The delta $\kron{i}{m}$ aligns the target query position $i$ with the probed position $m$, while $\kron{a}{j}$ sifts the sum over $a$ to $j$. We chain the upstream gradient:
  \begin{align}
    \dL{A_{mj}}
     & = \sum_{i<T} \sum_{d<h} \dL{Y_{id}}\,\pdv{Y_{id}}{A_{mj}} \notag \\
     & = \sum_{i<T} \sum_{d<h} \dL{Y_{id}}\,\kron{i}{m}\,V_{jd} \notag  \\
     & = \sum_{d<h} \dL{Y_{md}}\,V_{jd}
    = \Bigl( \dL{Y}\,V^{\top} \Bigr)_{mj}.
    \label{eq:att-chain-da}
  \end{align}
  Here, the delta $\kron{i}{m}$ sifts the sum over $i$ to $m$. The remaining sum contracts over the channel index $d$, representing the row of $\dL{Y}$ and the column of $V^{\top}$ (via $(V^{\top})_{dj} = V_{jd}$), which matches the matrix multiplication definition \eqref{eq:matprod} and yields $\dL{A} = \dL{Y}\,V^{\top}$.

  \paragraph{Backpropagation through row-wise Softmax}
  Each row of the attention matrix $A$ is an independent softmax over the corresponding row of raw scores $S$. Applying the softmax gradient result \eqref{eq:softmax-bwd} directly to each row $i$ yields:
  \begin{equation}
    \dL{S_{ij}} \;=\; A_{ij}\,\left( \dL{A_{ij}} - D_i \right),
    \label{eq:att-ds-local}
  \end{equation}
  where the row-wise dot product statistics term is defined as:
  \begin{equation}
    D_i \;=\; \sum_{k<T} \dL{A_{ik}}\,A_{ik}.
    \label{eq:att-di-def}
  \end{equation}
  Standard backpropagation materializes the full $T \times T$ gradient matrix $\dL{A}$ and computes the row-wise statistics $D_i$ before evaluating $\dL{S_{ij}}$.

  \paragraph{Local derivative $\pdv{S}{Q}$}
  With target score $S_{ij}$ and probe query $Q_{mk}$ (query position $m$, channel $k$):
  \begin{align}
    \pdv{S_{ij}}{Q_{mk}}
     & = \pdv{}{Q_{mk}} \left( \frac{1}{\sqrt{h}} \sum_{d<h} Q_{id}\,K_{jd} \right) \notag \\
     & = \frac{1}{\sqrt{h}} \sum_{d<h} \kron{i}{m}\,\kron{d}{k}\,K_{jd} \notag             \\
     & = \frac{1}{\sqrt{h}}\,\kron{i}{m}\,K_{jk},
    \label{eq:att-dsdq}
  \end{align}
  where the Kronecker deltas $\kron{i}{m}$ and $\kron{d}{k}$ enforce query position and channel alignment, sifting the channel sum to $d = k$. We chain the score gradient:
  \begin{align}
    \dL{Q_{mk}}
     & = \sum_{i<T} \sum_{j<T} \dL{S_{ij}}\,\pdv{S_{ij}}{Q_{mk}} \notag                    \\
     & = \sum_{i<T} \sum_{j<T} \dL{S_{ij}}\,\frac{1}{\sqrt{h}}\,\kron{i}{m}\,K_{jk} \notag \\
     & = \frac{1}{\sqrt{h}} \sum_{j<T} \dL{S_{mj}}\,K_{jk} \notag                          \\
     & = \left( \frac{1}{\sqrt{h}}\,\dL{S}\,K \right)_{mk},
    \label{eq:att-chain-dq}
  \end{align}
  where the query position delta sifts the outer sum over $i$ to $m$. The remaining sum over key position $j$ represents the inner product contraction between the rows of $\dL{S}$ and the columns of $K$, yielding $\dL{Q} = \frac{1}{\sqrt{h}}\,\dL{S}\,K$.

  \paragraph{Local derivative $\pdv{S}{K}$}
  With target score $S_{ij}$ and probe key $K_{mk}$ (key position $m$, channel $k$):
  \begin{align}
    \pdv{S_{ij}}{K_{mk}}
     & = \pdv{}{K_{mk}} \left( \frac{1}{\sqrt{h}} \sum_{d<h} Q_{id}\,K_{jd} \right) \notag \\
     & = \frac{1}{\sqrt{h}} \sum_{d<h} Q_{id}\,\kron{j}{m}\,\kron{d}{k} \notag             \\
     & = \frac{1}{\sqrt{h}}\,\kron{j}{m}\,Q_{ik},
    \label{eq:att-dsdk}
  \end{align}
  where the row delta $\kron{j}{m}$ indicates key position alignment, and the channel delta sifts the sum to $d = k$. We chain the score gradient:
  \begin{align}
    \dL{K_{mk}}
     & = \sum_{i<T} \sum_{j<T} \dL{S_{ij}}\,\pdv{S_{ij}}{K_{mk}} \notag                    \\
     & = \sum_{i<T} \sum_{j<T} \dL{S_{ij}}\,\frac{1}{\sqrt{h}}\,\kron{j}{m}\,Q_{ik} \notag \\
     & = \frac{1}{\sqrt{h}} \sum_{i<T} \dL{S_{im}}\,Q_{ik} \notag                          \\
     & = \frac{1}{\sqrt{h}} \sum_{i<T} (\dL{S}^{\top})_{mi}\,Q_{ik} \notag                 \\
     & = \left( \frac{1}{\sqrt{h}}\,\dL{S}^{\top}\,Q \right)_{mk},
    \label{eq:att-chain-dk}
  \end{align}
  where the key position delta sifts the inner sum over $j$ down to $j = m$. The remaining sum over query position $i$ contracts the column dimension of $\dL{S}$ with the row dimension of $Q$. To express this as a matrix multiplication, we transpose $\dL{S}$ so that the summed index $i$ sits as the column index of the left factor, which yields the matrix product $\dL{K} = \frac{1}{\sqrt{h}}\,\dL{S}^{\top}\,Q$.

  \paragraph{FlashAttention-2 Simplification: Softmax Statistics Contraction}
  To avoid materializing the intermediate $T \times T$ matrices $\dL{A}$ and $\dL{S}$ in DRAM, FlashAttention-2 \cite{flashattention2} simplifies the softmax statistics term $D_i$. We substitute the definition of $\dL{A_{ik}} = \sum_{d<h} \dL{Y_{id}}\,V_{kd}$ from \eqref{eq:att-chain-da} into the definition of $D_i$ \eqref{eq:att-di-def} to obtain:
  \begin{align}
    D_i
     & = \sum_{k<T} \left( \sum_{d<h} \dL{Y_{id}}\,V_{kd} \right)\,A_{ik} \notag \\
     & = \sum_{d<h} \dL{Y_{id}}\,\left( \sum_{k<T} A_{ik}\,V_{kd} \right) \notag \\
     & = \sum_{d<h} \dL{Y_{id}}\,Y_{id},
    \label{eq:att-di-contract}
  \end{align}
  where the forward pass output $Y_{id}$ replaces the sum over key positions $k$ in the second line via \eqref{eq:att-output-fwd}. This contraction yields a key mathematical optimization: the row statistics term $D_i$ is the row-wise dot product of the output $Y$ and its gradient $\dL{Y}$. Precomputing $D_i \in \R^T$ directly from the output tensors allows us to evaluate the score gradients $\dL{S_{ij}}$ block-by-block in SRAM:
  \begin{equation}
    \dL{S_{ij}} \;=\; A_{ij}\,\left( \sum_{d<h} \dL{Y_{id}}\,V_{jd} - D_i \right),
    \label{eq:att-ds-final}
  \end{equation}
  eliminating the need to write the large $\dL{A}$ matrix to DRAM.

  \paragraph{FlashAttention-2 Memory-Efficient Block Tiling}
  Standard backpropagation materializes the $T \times T$ matrices $S$, $A$, $\dL{A}$, and $\dL{S}$ in DRAM, which creates a massive $O(T^2)$ memory footprint. FlashAttention-2 avoids this bottleneck by tiling the sequence dimension and executing the backward pass in SRAM in two passes:
  \begin{enumerate}
    \item \textbf{Pass 1 ($dQ$ Computation):} We tile query positions into blocks of size $B_r$ (indexing rows $i$). For each block, we:
          \begin{itemize}
            \item load $Q_i$, $\dL{Y}_i$, $Y_i$, and $L_i$ (the LSE vector) from DRAM to SRAM;
            \item compute the row scaling term $D_i = \sum_{d<h} \dL{Y}_{id}\,Y_{id}$ in SRAM;
            \item loop over key/value blocks $j$ of size $B_c$: load $K_j, V_j$ to SRAM, compute the score block $S_{ij} = Q_i K_j^{\top}/\sqrt{h}$, mask elements where $j > i$, apply softmax $A_{ij} = \exp(S_{ij} - L_i)$, compute the gradient slice $\dL{S_{ij}} = A_{ij} (\dL{Y}_i V_j^{\top} - D_i)$, and accumulate:
                  \begin{equation}
                    \dL{Q_i} \;\mathrel{+}=\; \frac{1}{\sqrt{h}}\,\dL{S_{ij}}\,K_j.
                  \end{equation}
            \item write the fully accumulated $\dL{Q_i}$ to DRAM once the loop over $j$ completes.
          \end{itemize}
    \item \textbf{Pass 2 ($dK$ and $dV$ Computation):} We tile key and value positions into blocks of size $B_c$ (indexing columns $j$). For each block, we:
          \begin{itemize}
            \item load $K_j, V_j$ to SRAM and initialize local accumulators $\dL{K_j} = 0, \dL{V_j} = 0$;
            \item loop over query blocks $i$ of size $B_r$: load $Q_i$, $\dL{Y}_i$, $Y_i$, and $L_i$, compute $D_i$, reconstruct the local slices $A_{ij}$ and $\dL{S_{ij}}$, and accumulate:
                  \begin{align}
                    \dL{V_j} \;\mathrel{+} & =\; A_{ij}^{\top}\,\dL{Y}_i,                     \\
                    \dL{K_j} \;\mathrel{+} & =\; \frac{1}{\sqrt{h}}\,\dL{S_{ij}}^{\top}\,Q_i.
                  \end{align}
            \item write the fully accumulated $\dL{K_j}$ and $\dL{V_j}$ to DRAM once the loop over $i$ completes.
          \end{itemize}
  \end{enumerate}
  Both passes are contention-free (Pass 1 parallelizes over $i$, while Pass 2 parallelizes over $j$) and execute without global atomic operations.
\end{derivation}

\begin{result}
  For each head, the gradients w.r.t.\ Queries $Q$, Keys $K$, and Values $V$ are:
  \begin{align}
    \dL{Q} \; & =\; \frac{1}{\sqrt{h}}\,\dL{S}\,K,        \\
    \dL{K} \; & =\; \frac{1}{\sqrt{h}}\,\dL{S}^{\top}\,Q, \\
    \dL{V} \; & =\; A^{\top}\,\dL{Y},
  \end{align}
  where the causal-masked score gradient $\dL{S} \in \R^{T \times T}$ is:
  \begin{equation}
    \dL{S_{ij}} \;=\; \begin{cases}
      A_{ij} \left( \dL{A_{ij}} - D_i \right) & j \leq i, \\
      0                                       & j > i,
    \end{cases}
  \end{equation}
  with the intermediate term and scaling statistics:
  \begin{align}
    \dL{A_{ij}} \; & =\; \sum_{d<h} \dL{Y_{id}}\,V_{jd}, \\
    D_i \;         & =\; \sum_{d<h} \dL{Y_{id}}\,Y_{id}.
  \end{align}
\end{result}

% =============================================================================
\section{Embeddings (Encoder)}\label{sec:encoder}

The input representation encodes both token vocabulary identity and sequence position. Following the standard language modeling paradigm \cite{embeddings, attention}, the forward pass computes the input sequence embeddings $x_{bt}$ by adding token and position embeddings, or in index form:
\begin{equation}
  x_{bt, c} \;=\; W_{E,\, \mathrm{tok}_{bt},\, c} + W_{P,\, t,\, c},
  \label{eq:enc-fwd}
\end{equation}
where token indices satisfy $\mathrm{tok}_{bt} \in \{0, \dots, V-1\}$, token embeddings occupy $W_E \in \R^{V \times C}$, position embeddings occupy $W_P \in \R^{T \times C}$, and the channel index is $c < C$.

\begin{derivation}
  Let $b$, $t$, and $c$ denote the batch, sequence position, and channel indices. We introduce independent probe indices $k < T$ (position), $v < V$ (vocab token), and $j < C$ (channel).

  \paragraph{Gradient w.r.t.\ position embeddings}
  To compute the gradient w.r.t.\ position embeddings, we probe $W_{P,\, k,\, j}$ (position $k$, channel $j$). Because the token lookup remains constant, we differentiate the position term using \eqref{eq:entry-dv}:
  \begin{equation}
    \pdv{W_{P,\, t,\, c}}{W_{P,\, k,\, j}} \;=\; \kron{t}{k}\,\kron{c}{j},
    \label{eq:enc-dxdwp-local}
  \end{equation}
  which yields the local derivative:
  \begin{equation}
    \pdv{x_{bt, c}}{W_{P,\, k,\, j}} \;=\; \kron{t}{k}\,\kron{c}{j}.
    \label{eq:enc-dxdwp}
  \end{equation}
  We chain the upstream gradient $\dL{x_{bt, c}}$ using \eqref{eq:chain}:
  \begin{align}
    \dL{W_{P,\, k,\, j}}
     & = \sum_{b<B} \sum_{t<T} \sum_{c<C} \dL{x_{bt, c}}\,\pdv{x_{bt, c}}{W_{P,\, k,\, j}} \notag \\
     & = \sum_{b<B} \sum_{t<T} \sum_{c<C} \dL{x_{bt, c}}\,\kron{t}{k}\,\kron{c}{j}.
  \end{align}
  The Kronecker deltas act as switches that vanish for all terms except where the indices match: $\kron{t}{k}$ is non-zero only when $t = k$, and $\kron{c}{j}$ is non-zero only when $c = j$. These deltas sift the summation, collapsing the double sum over position $t$ and channel $c$ to a single term where $t$ is replaced by $k$ and $c$ is replaced by $j$. The sum over the batch dimension $b$ remains because the positional embeddings $W_P$ are shared across all batches, meaning they receive gradient updates from every batch element:
  \begin{equation}
    \dL{W_{P,\, k,\, j}} \;=\; \sum_{b<B} \dL{x_{bk, j}},
    \label{eq:enc-wp-bwd-idx}
  \end{equation}
  with free indices $(k, j)$. Because the network shares $W_{P, t}$ across batches, the gradient reduces over the batch dimension $b$.

  \paragraph{Gradient w.r.t.\ token embeddings}
  To compute the gradient w.r.t.\ token embeddings, we probe $W_{E,\, v,\, j}$ (vocab token $v$, channel $j$), which requires comparing the lookup index $\mathrm{tok}_{bt}$ against $v$. We write the embedding lookup in indicator form:
  \begin{equation}
    W_{E,\, \mathrm{tok}_{bt},\, c} \;=\; \sum_{u<V} \ind{\mathrm{tok}_{bt} = u}\,W_{E,\, u,\, c}.
  \end{equation}
  We differentiate with respect to $W_{E,\, v,\, j}$:
  \begin{align}
    \pdv{W_{E,\, \mathrm{tok}_{bt},\, c}}{W_{E,\, v,\, j}}
     & = \sum_{u<V} \ind{\mathrm{tok}_{bt} = u}\,\pdv{W_{E,\, u,\, c}}{W_{E,\, v,\, j}} \notag \\
     & = \sum_{u<V} \ind{\mathrm{tok}_{bt} = u}\,\kron{u}{v}\,\kron{c}{j} \notag               \\
     & = \ind{\mathrm{tok}_{bt} = v}\,\kron{c}{j},
    \label{eq:enc-dxdwe-local}
  \end{align}
  which yields the local derivative:
  \begin{equation}
    \pdv{x_{bt, c}}{W_{E,\, v,\, j}} \;=\; \ind{\mathrm{tok}_{bt} = v}\,\kron{c}{j}.
    \label{eq:enc-dxdwe}
  \end{equation}
  We chain the upstream gradient $\dL{x_{bt, c}}$ using \eqref{eq:chain}:
  \begin{align}
    \dL{W_{E,\, v,\, j}}
     & = \sum_{b<B} \sum_{t<T} \sum_{c<C} \dL{x_{bt, c}}\,\pdv{x_{bt, c}}{W_{E,\, v,\, j}} \notag   \\
     & = \sum_{b<B} \sum_{t<T} \sum_{c<C} \dL{x_{bt, c}}\,\ind{\mathrm{tok}_{bt} = v}\,\kron{c}{j}.
  \end{align}
  The delta $\kron{c}{j}$ sifts the sum over channel $c$ to $j$. The indicator function $\ind{\mathrm{tok}_{bt} = v}$ evaluates to $1$ when the active token at position $t$ in batch $b$ equals the vocab index $v$, and $0$ otherwise. When we chain the upstream gradient, this indicator acts as a filter on the remaining sums over batch $b$ and position $t$, accumulating gradients only from the specific sequence positions where token $v$ was present:
  \begin{equation}
    \dL{W_{E,\, v,\, j}} \;=\; \sum_{b<B} \sum_{t<T} \dL{x_{bt, j}}\,\ind{\mathrm{tok}_{bt} = v},
    \label{eq:enc-we-bwd-idx}
  \end{equation}
  with free indices $(v, j)$. This operation scatter-accumulates upstream gradients at matching token locations, following the $\mathrel{+}=$ convention of \S\ref{sec:notation} (often implemented as an index-add or scatter-add operation in deep learning libraries).
\end{derivation}

\begin{result}
  The token and positional embedding gradients are:
  \begin{align}
    \dL{W_{E,\, v,\, c}} \; & =\; \sum_{b<B} \sum_{t<T} \dL{x_{bt, c}}\,\ind{\mathrm{tok}_{bt} = v}, \label{eq:enc-we-bwd} \\
    \dL{W_{P,\, t,\, c}} \; & =\; \sum_{b<B} \dL{x_{bt, c}}. \label{eq:enc-wp-bwd}
  \end{align}
  The positional embedding gradient $\dL{W_P}$ reduces the upstream gradient over the batch dimension, while the token embedding gradient $\dL{W_E}$ accumulates contributions via scatter-add at active token locations.
\end{result}

% =============================================================================
\section{The Full Backward Pass}\label{sec:full}

We trace the complete backward pass in reverse order of forward execution. Gradients flow from the cross-entropy loss, back through the language modeling head, through the stack of $N_{\text{layers}}$ transformer blocks, and finally accumulate in the token and position embeddings. We write out the complete, coordinate-level gradient chain in index form.

\subsection{Top-level Gradient Flow}

We define the total loss $L$ as the mean over all batch elements and sequence positions, $L = \frac{1}{B \cdot T} \sum_{b<B} \sum_{t<T} L_{bt}$. The backward pass begins by differentiating $L$ w.r.t.\ the logits $Z_{bt, v}$ at position $t$ in batch $b$ and vocabulary entry $v < V_p$:
\begin{equation}
  \dL{Z_{bt,\, v}} \;=\; \frac{1}{B \cdot T} \left( p_{bt, v} - \ind{v = y_{bt}} \right).
  \label{eq:full-dlogits}
\end{equation}

\paragraph{LM Head Layer}
The logits are computed from the final LayerNorm output $X_{\text{final}, bt, c}$ and LM head weights $W_{\text{LM}, v, c}$ as $Z_{bt, v} = \sum_{c<C} X_{\text{final}, bt, c} W_{\text{LM}, v, c}$. We apply the coordinate-level chain rule to propagate the logit gradient:
\begin{align}
  \dL{X_{\text{final},\, bt,\, c}}
   & = \sum_{v<V_p} \dL{Z_{bt,\, v}}\,W_{\text{LM},\, v,\, c},
  \label{eq:full-dxfinal}                                                   \\
  \dL{W_{\text{LM},\, v,\, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{Z_{bt,\, v}}\,X_{\text{final},\, bt,\, c},
  \label{eq:full-dwlm}
\end{align}

\paragraph{Final LayerNorm}
Let $x^{(N)} \in \R^{B \times T \times C}$ denote the output of the last transformer block. We normalize each row to compute $X_{\text{final}, bt, c} = \gamma_{f, c} \hat{x}^{(N)}_{bt, c} + \beta_{f, c}$. We chain the final LayerNorm gradient back to the block output:
\begin{align}
  \dL{\gamma_{f,\, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{X_{\text{final},\, bt,\, c}}\,\hat{x}^{(N)}_{bt,\, c},
  \label{eq:full-dgamma-f}                                                              \\
  \dL{\beta_{f,\, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{X_{\text{final},\, bt,\, c}},
  \label{eq:full-dbeta-f}
\end{align}
and the input gradient:
\begin{align}
  \dL{x^{(N)}_{bt,\, c}}
   & = \frac{1}{\sqrt{\sigma^2_{bt}}} \left( \gamma_{f, c}\,\dL{X_{\text{final},\, bt,\, c}} \right. \notag                                        \\
   & \quad - \frac{1}{C} \sum_{k<C} \gamma_{f, k}\,\dL{X_{\text{final},\, bt,\, k}} \notag                                                         \\
   & \quad \left. - \frac{\hat{x}^{(N)}_{bt,\, c}}{C} \sum_{k<C} \gamma_{f, k}\,\dL{X_{\text{final},\, bt,\, k}}\,\hat{x}^{(N)}_{bt,\, k} \right),
  \label{eq:full-dxn}
\end{align}
where $\sigma^2_{bt}$ is the variance and $\hat{x}^{(N)}_{bt, c}$ is the normalized value of row $x^{(N)}_{bt}$ computed during the forward pass.

\subsection{The Transformer Block Stack}

For layer $l = N - 1$ down to $0$, the block maps the input $x^{(l)}$ to $x^{(l+1)}$ using residual connections \cite{resnet} containing multi-head self-attention \cite{attention} and a multi-layer perceptron:
\begin{align}
  x^{\text{mid}, (l)}_{bt,\, c} \; & =\; x^{(l)}_{bt,\, c} + z^{\text{attn}, (l)}_{bt,\, c},            \\
  x^{(l+1)}_{bt,\, c} \;           & =\; x^{\text{mid}, (l)}_{bt,\, c} + z^{\text{mlp}, (l)}_{bt,\, c},
\end{align}
where $z^{\text{attn}, (l)}$ and $z^{\text{mlp}, (l)}$ are the attention and MLP outputs. We backpropagate the block output gradient $\dL{x^{(l+1)}_{bt, c}}$ in two steps.

\paragraph{Step One: The MLP Sub-Layer}
Due to the residual connection, the upstream gradient w.r.t.\ $z^{\text{mlp}, (l)}$ is $\dL{z^{\text{mlp}, (l)}_{bt,\, c}} = \dL{x^{(l+1)}_{bt,\, c}}$. The MLP sub-layer forward pass evaluates:
\begin{align}
  a^{(l)}_{bt,\, c} \;              & =\; \gamma^{(l)}_{2, c}\,\hat{x}^{\text{mid}, (l)}_{bt,\, c} + \beta^{(l)}_{2, c},                        \\
  h^{\text{fc}, (l)}_{bt,\, i} \;   & =\; \sum_{c<C} a^{(l)}_{bt,\, c}\,W^{\text{fc}, (l)}_{i,\, c} + b^{\text{fc}, (l)}_{i},                   \\
  h^{\text{gelu}, (l)}_{bt,\, i} \; & =\; \gelu\left( h^{\text{fc}, (l)}_{bt,\, i} \right),                                                     \\
  z^{\text{mlp}, (l)}_{bt,\, c} \;  & =\; \sum_{i<4C} h^{\text{gelu}, (l)}_{bt,\, i}\,W^{\text{proj}, (l)}_{c,\, i} + b^{\text{proj}, (l)}_{c},
\end{align}
for $i < 4C$. We differentiate in reverse order. First, for the projection layer, we contract over the output channel index $c < C$:
\begin{align}
  \dL{h^{\text{gelu}, (l)}_{bt,\, i}}
   & = \sum_{c<C} \dL{z^{\text{mlp}, (l)}_{bt,\, c}}\,W^{\text{proj}, (l)}_{c,\, i},
  \label{eq:full-dgelu}                                                                          \\
  \dL{W^{\text{proj}, (l)}_{c,\, i}}
   & = \sum_{b<B} \sum_{t<T} \dL{z^{\text{mlp}, (l)}_{bt,\, c}}\,h^{\text{gelu}, (l)}_{bt,\, i},
  \label{eq:full-dwproj}                                                                         \\
  \dL{b^{\text{proj}, (l)}_{c}}
   & = \sum_{b<B} \sum_{t<T} \dL{z^{\text{mlp}, (l)}_{bt,\, c}}.
  \label{eq:full-dbproj}
\end{align}
Second, the GELU derivative sifts the activation gradient pointwise:
\begin{equation}
  \dL{h^{\text{fc}, (l)}_{bt,\, i}} \;=\; \dL{h^{\text{gelu}, (l)}_{bt,\, i}}\,\gelu'\left( h^{\text{fc}, (l)}_{bt,\, i} \right).
  \label{eq:full-dfc}
\end{equation}
Third, the first linear layer contracts over the hidden dimension $i < 4C$:
\begin{align}
  \dL{a^{(l)}_{bt,\, c}}
   & = \sum_{i<4C} \dL{h^{\text{fc}, (l)}_{bt,\, i}}\,W^{\text{fc}, (l)}_{i,\, c},
  \label{eq:full-da}                                                               \\
  \dL{W^{\text{fc}, (l)}_{i,\, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{h^{\text{fc}, (l)}_{bt,\, i}}\,a^{(l)}_{bt,\, c},
  \label{eq:full-dwfc}                                                             \\
  \dL{b^{\text{fc}, (l)}_{i}}
   & = \sum_{b<B} \sum_{t<T} \dL{h^{\text{fc}, (l)}_{bt,\, i}}.
  \label{eq:full-dbfc}
\end{align}
Fourth, we backpropagate through LayerNorm 2 using stats $\mu^{(2)}_{bt}$ and $\sigma^{2, (2)}_{bt}$:
\begin{align}
  \dL{\gamma^{(l)}_{2, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{a^{(l)}_{bt,\, c}}\,\hat{x}^{\text{mid}, (l)}_{bt,\, c},
  \label{eq:full-dgamma2}                                                                 \\
  \dL{\beta^{(l)}_{2, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{a^{(l)}_{bt,\, c}},
  \label{eq:full-dbeta2}
\end{align}
and the input gradient:
\begin{align}
  \dL{x^{\text{mid}, (l)}_{bt,\, c}\Big|_{\text{MLP}}}
   & = \frac{1}{\sqrt{\sigma^{2, (2)}_{bt}}} \left( \gamma^{(l)}_{2, c}\,\dL{a^{(l)}_{bt,\, c}} \right. \notag                                                         \\
   & \quad - \frac{1}{C} \sum_{k<C} \gamma^{(l)}_{2, k}\,\dL{a^{(l)}_{bt,\, k}} \notag                                                                                 \\
   & \quad \left. - \frac{\hat{x}^{\text{mid}, (l)}_{bt,\, c}}{C} \sum_{k<C} \gamma^{(l)}_{2, k}\,\dL{a^{(l)}_{bt,\, k}}\,\hat{x}^{\text{mid}, (l)}_{bt,\, k} \right).
  \label{eq:full-dxmid-mlp}
\end{align}
The total mid-state gradient accumulates both pathways:
\begin{equation}
  \dL{x^{\text{mid}, (l)}_{bt,\, c}} \;=\; \dL{x^{(l+1)}_{bt,\, c}} + \dL{x^{\text{mid}, (l)}_{bt,\, c}\Big|_{\text{MLP}}}.
  \label{eq:full-dxmid}
\end{equation}

\paragraph{Step Two: The Attention Sub-Layer}
Due to the residual connection, the upstream gradient w.r.t.\ $z^{\text{attn}, (l)}$ is $\dL{z^{\text{attn}, (l)}_{bt,\, c}} = \dL{x^{\text{mid}, (l)}_{bt,\, c}}$. The attention sub-layer forward pass evaluates:
\begin{align}
  h^{\text{attn\_inp}, (l)}_{bt,\, c} \; & =\; \gamma^{(l)}_{1, c}\,\hat{x}^{(l)}_{bt,\, c} + \beta^{(l)}_{1, c},                                        \\
  qkv^{(l)}_{bt,\, m} \;                 & =\; \sum_{c<C} h^{\text{attn\_inp}, (l)}_{bt,\, c}\,W^{\text{attn}, (l)}_{m,\, c} + b^{\text{attn}, (l)}_{m}, \\
  z^{\text{attn}, (l)}_{bt,\, c} \;      & =\; \sum_{k<C} Y^{\text{heads}, (l)}_{bt,\, k}\,W^{\text{proj}, (l)}_{c,\, k} + b^{\text{proj}, (l)}_{c},
\end{align}
for $m < 3C$. First, we propagate gradients through the output projection:
\begin{align}
  \dL{Y^{\text{heads}, (l)}_{bt,\, k}}
   & = \sum_{c<C} \dL{z^{\text{attn}, (l)}_{bt,\, c}}\,W^{\text{proj}, (l)}_{c,\, k},
  \label{eq:full-dyheads}                                                                          \\
  \dL{W^{\text{proj}, (l)}_{c,\, k}}
   & = \sum_{b<B} \sum_{t<T} \dL{z^{\text{attn}, (l)}_{bt,\, c}}\,Y^{\text{heads}, (l)}_{bt,\, k},
  \label{eq:full-dwproj-attn}                                                                      \\
  \dL{b^{\text{proj}, (l)}_{c}}
   & = \sum_{b<B} \sum_{t<T} \dL{z^{\text{attn}, (l)}_{bt,\, c}}.
  \label{eq:full-dbproj-attn}
\end{align}
Second, we partition the gradient w.r.t.\ the head activations. Let $H$ be the head count, $h = C/H$ be the head dimension, $i < H$ be the head index, and $d < h$ be the head channel index. We extract the head output gradient:
\begin{equation}
  \dL{Y^{(l)}_{bt,\, i,\, d}} \;=\; \dL{Y^{\text{heads}, (l)}_{bt,\, i \cdot h + d}}.
\end{equation}
We backpropagate through the self-attention mechanism for each head $i$:
\begin{align}
  \dL{V^{(l)}_{bj,\, i,\, d}}
   & = \sum_{t<T} \dL{Y^{(l)}_{bt,\, i,\, d}}\,A^{(l)}_{b,\, i,\, t,\, j},
  \label{eq:full-dv}                                                       \\
  \dL{A^{(l)}_{b,\, i,\, t,\, j}}
   & = \sum_{d<h} \dL{Y^{(l)}_{bt,\, i,\, d}}\,V^{(l)}_{bj,\, i,\, d}.
  \label{eq:full-da-matrix}
\end{align}
The row-wise softmax statistics $D^{(l)}_{b,\, i,\, t}$ contract to:
\begin{equation}
  D^{(l)}_{b,\, i,\, t} \;=\; \sum_{d<h} \dL{Y^{(l)}_{bt,\, i,\, d}}\,Y^{(l)}_{bt,\, i,\, d}.
  \label{eq:full-dstats-attn}
\end{equation}
The score gradient $\dL{S^{(l)}_{b,\, i,\, t,\, j}}$ is masked causally:
\begin{equation}
  \dL{S^{(l)}_{b,\, i,\, t,\, j}} \;=\; \begin{cases}
    A^{(l)}_{b,\, i,\, t,\, j}\,\left( \dL{A^{(l)}_{b,\, i,\, t,\, j}} - D^{(l)}_{b,\, i,\, t} \right) & j \leq t, \\
    0                                                                                                  & j > t.
  \end{cases}
  \label{eq:full-ds}
\end{equation}
We compute query and key gradients by contracting over the sequence dimensions:
\begin{align}
  \dL{Q^{(l)}_{bt,\, i,\, d}}
   & = \frac{1}{\sqrt{h}} \sum_{j<T} \dL{S^{(l)}_{b,\, i,\, t,\, j}}\,K^{(l)}_{bj,\, i,\, d},
  \label{eq:full-dq}                                                                          \\
  \dL{K^{(l)}_{bj,\, i,\, d}}
   & = \frac{1}{\sqrt{h}} \sum_{t<T} \dL{S^{(l)}_{b,\, i,\, t,\, j}}\,Q^{(l)}_{bt,\, i,\, d}.
  \label{eq:full-dk}
\end{align}
Third, we reassemble the composite projection gradient $\dL{qkv^{(l)}_{bt,\, m}}$:
\begin{equation}
  \dL{qkv^{(l)}_{bt,\, m}} \;=\; \begin{cases}
    \dL{Q^{(l)}_{bt,\, i,\, d}} & m = i \cdot h + d,      \\
    \dL{K^{(l)}_{bt,\, i,\, d}} & m = C + i \cdot h + d,  \\
    \dL{V^{(l)}_{bt,\, i,\, d}} & m = 2C + i \cdot h + d,
  \end{cases}
  \label{eq:full-dqkv-reassemble}
\end{equation}
where $i = \lfloor (m \bmod C) / h \rfloor$ and $d = m \bmod h$.
Fourth, we backpropagate through the QKV projection layer parameters and inputs:
\begin{align}
  \dL{h^{\text{attn\_inp}, (l)}_{bt,\, c}}
   & = \sum_{m<3C} \dL{qkv^{(l)}_{bt,\, m}}\,W^{\text{attn}, (l)}_{m,\, c},
  \label{eq:full-dhattninp}                                                                 \\
  \dL{W^{\text{attn}, (l)}_{m,\, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{qkv^{(l)}_{bt,\, m}}\,h^{\text{attn\_inp}, (l)}_{bt,\, c},
  \label{eq:full-dwattn}                                                                    \\
  \dL{b^{\text{attn}, (l)}_{m}}
   & = \sum_{b<B} \sum_{t<T} \dL{qkv^{(l)}_{bt,\, m}}.
  \label{eq:full-dbattn}
\end{align}
Fifth, we backpropagate through LayerNorm 1 using stats $\mu^{(1)}_{bt}$ and $\sigma^{2, (1)}_{bt}$:
\begin{align}
  \dL{\gamma^{(l)}_{1, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{h^{\text{attn\_inp}, (l)}_{bt,\, c}}\,\hat{x}^{(l)}_{bt,\, c},
  \label{eq:full-dgamma1}                                                                       \\
  \dL{\beta^{(l)}_{1, c}}
   & = \sum_{b<B} \sum_{t<T} \dL{h^{\text{attn\_inp}, (l)}_{bt,\, c}},
  \label{eq:full-dbeta1}
\end{align}
and the input gradient through the attention pathway:
\begin{align}
  \dL{x^{(l)}_{bt,\, c}\Big|_{\text{attn}}}
   & = \frac{1}{\sqrt{\sigma^{2, (1)}_{bt}}} \left( \gamma^{(l)}_{1, c}\,\dL{h^{\text{attn\_inp}, (l)}_{bt,\, c}} \right. \notag                                 \\
   & \quad - \frac{1}{C} \sum_{k<C} \gamma^{(l)}_{1, k}\,\dL{h^{\text{attn\_inp}, (l)}_{bt,\, k}} \notag                                                         \\
   & \quad \left. - \frac{\hat{x}^{(l)}_{bt,\, c}}{C} \sum_{k<C} \gamma^{(l)}_{1, k}\,\dL{h^{\text{attn\_inp}, (l)}_{bt,\, k}}\,\hat{x}^{(l)}_{bt,\, k} \right).
  \label{eq:full-dx-attn}
\end{align}
The total block input gradient accumulates both pathways:
\begin{equation}
  \dL{x^{(l)}_{bt,\, c}} \;=\; \dL{x^{\text{mid}, (l)}_{bt,\, c}} + \dL{x^{(l)}_{bt,\, c}\Big|_{\text{attn}}}.
  \label{eq:full-dx}
\end{equation}

\subsection{Parameter Accumulation and Weight Tying}

At the bottom of the stack, the gradient w.r.t.\ the input embeddings is $\dL{x^{(0)}_{bt,\, c}}$. The forward embedding layer adds token and position embeddings, $x^{(0)}_{bt, c} = W_{E, \text{tok}_{bt}, c} + W_{P, t, c}$.
Differentiating w.r.t.\ position embeddings $W_{P, t, c}$ reduces the gradient over the batch dimension $b$:
\begin{equation}
  \dL{W_{P,\, t,\, c}} \;=\; \sum_{b<B} \dL{x^{(0)}_{bt,\, c}}.
  \label{eq:full-dwp}
\end{equation}
Differentiating w.r.t.\ token embeddings $W_{E, v, c}$ scatter-accumulates using the indicator function:
\begin{equation}
  \dL{W_{E,\, v,\, c}} \;=\; \sum_{b<B} \sum_{t<T} \dL{x^{(0)}_{bt,\, c}}\,\ind{\text{tok}_{bt} = v}.
  \label{eq:full-dwe}
\end{equation}
Because the input token embeddings $W_E$ and LM head weights $W_{\text{LM}}$ tie ($W_E = W_{\text{LM}}$), their gradients sum. The total tied token embedding gradient is:
\begin{align}
  \dL{W_{E,\, v,\, c}}_{\text{tied}}
   & = \sum_{b<B} \sum_{t<T} \left( \dL{x^{(0)}_{bt,\, c}}\,\ind{\text{tok}_{bt} = v} \right. \notag \\
   & \quad \left. + \, \dL{Z_{bt,\, v}}\,X_{\text{final},\, bt,\, c} \right),
  \label{eq:full-dwetied-expand}
\end{align}
which accumulates contributions from the embedding lookup at the network base and the output projection at the network top.

\section*{AI Use Statement}
In writing these derivations, we used the following AI tools: Overleaf's spell check and grammar modules to correct spelling and grammar mistakes; Cursor's Code and \LaTeX\ plugins to help with formatting and outlining sections with \LaTeX\ autocompletion; and Cursor's Composer 2.5 for bulk \LaTeX\ fixes and resolving compiler errors. For the mojo code we are not using any AI tools or LSPs.

% =============================================================================
% =============================================================================
\appendices
\section{Gradient Checking}\label{app:gradcheck}

We use a central finite difference approximation to check the numerical accuracy of each derivation:
\begin{equation}
  \pdv{L}{\theta_i} \;\approx\;
  \frac{L(\theta + \epsilon e_i) - L(\theta - \epsilon e_i)}{2\epsilon}.
\end{equation}
Our verification tests enforce the relative (\texttt{rtol}) and absolute (\texttt{atol}) tolerances defined in \texttt{tests/\_dtypes.py}, as summarized in Table~\ref{tab:tolerances}.

\begin{table}[h]
  \centering
  \caption{Per-dtype tolerances for PyTorch reference comparison.}
  \label{tab:tolerances}
  \begin{tabular}{@{}lll@{}}
    \toprule
    Dtype             & Absolute Tolerance (\texttt{atol}) & Relative Tolerance (\texttt{rtol}) \\
    \midrule
    \texttt{float32}  & $10^{-6}$                          & $10^{-5}$                          \\
    \texttt{float16}  & $10^{-3}$                          & $10^{-2}$                          \\
    \texttt{bfloat16} & $5 \times 10^{-3}$                 & $2 \times 10^{-2}$                 \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Useful Identities}\label{app:identities}

\begin{itemize}
  \item Derivatives of basic functions:
        \begin{align*}
          \dv{}{x}\log x \;  & =\; \frac{1}{x}, \qquad \dv{}{x} e^{x} \;=\; e^{x}, \\
          \dv{}{x}\tanh x \; & =\; 1 - \tanh^2 x.
        \end{align*}
  \item Quotient and product rules:
        \begin{equation*}
          \dv{}{x}\left(\frac{u}{v}\right) = \frac{u'v - uv'}{v^2}, \quad
          \dv{}{x}(uv) = u'v + uv'.
        \end{equation*}
  \item Kronecker delta sifting property:
        \begin{equation*}
          \sum_j \kron{j}{l}\,A_j \;=\; A_l.
        \end{equation*}
  \item Log-sum-exp function and its gradient:
        \begin{equation*}
          \pdv{}{x_i} \log \sum_j e^{x_j} \;=\; \frac{e^{x_i}}{\sum_j e^{x_j}} \;=\; \softmax(x)_i.
        \end{equation*}
\end{itemize}

\begin{thebibliography}{1}
  \bibitem{llmc}
  A.~Karpathy, ``llm.c: LLM training in simple, raw C/CUDA,'' 2024. [Online]. Available: \url{https://github.com/karpathy/llm.c}

  \bibitem{flashattention2}
  T.~Dao, ``FlashAttention-2: Faster attention with better parallelism and work partitioning,'' \emph{arXiv preprint arXiv:2307.08691}, 2023. [Online]. Available: \url{https://arxiv.org/abs/2307.08691}

  \bibitem{gelu}
  D.~Hendrycks and K.~Gimpel, ``Gaussian Error Linear Units (GELUs),'' \emph{arXiv preprint arXiv:1606.08415}, 2016. [Online]. Available: \url{https://arxiv.org/abs/1606.08415}

  \bibitem{page}
  E.~Page, ``Approximations to the Cumulative Normal Function and its Inverse for Use on a Pocket Calculator,'' \emph{Journal of the Royal Statistical Society. Series C (Applied Statistics)}, vol. 26, no. 1, pp. 75--76, 1977.

  \bibitem{layernorm}
  J.~L.~Ba, J.~R.~Kiros, and G.~E.~Hinton, ``Layer Normalization,'' \emph{arXiv preprint arXiv:1607.06450}, 2016. [Online]. Available: \url{https://arxiv.org/abs/1607.06450}

  \bibitem{embeddings}
  Y.~Bengio, R.~Ducharme, P.~Vincent, and C.~Jauvin, ``A Neural Probabilistic Language Model,'' \emph{Journal of Machine Learning Research}, vol. 3, pp. 1137--1155, 2003.

  \bibitem{attention}
  A.~Vaswani, N.~Shazeer, N.~Parmar, J.~Uszkoreit, L.~Jones, A.~N.~Gomez, L.~Kaiser, and I.~Polosukhin, ``Attention Is All You Need,'' in \emph{Advances in Neural Information Processing Systems}, 2017, pp. 5998--6008.

  \bibitem{resnet}
  K.~He, X.~Zhang, S.~Ren, and J.~Sun, ``Deep Residual Learning for Image Recognition,'' in \emph{IEEE Conference on Computer Vision and Pattern Recognition}, 2016, pp. 770--778.

\end{thebibliography}

\end{document}
